% =============================================================
% Appendix F: The infinite Pair HMM and its MCMC sampler
% Source: ~/tkf-dp/main.tex Section 7 (sec:f2-scfg + sec:aug-phmm +
% sec:infinite-hmm), quoted near-verbatim, plus the corresponding
% algorithmic spec in ~/tkf-dp/analysis/mcmc_infinite_phmm.md.
% =============================================================

The Pair HMM postprocessing methods of \S\ref{sec:postprocessing}
are progressively faithful approximations to a single underlying
object: the alignment-and-partition posterior on
$\pi(A, \pi_{\!M} \mid X, Y)$, where $A$ is the TKF92 alignment and
$\pi_{\!M}$ is a Chinese-restaurant-process partition of its Match
cells into singletons (block of size 1, contributing the standard
per-site CTMC factor) and pairs (block of size 2, contributing the
joint Potts CTMC factor). This appendix develops the three-factor
factorisation $\pi_{\TKF92} \cdot \pi_{\mathrm{CRP}} \cdot
\prod_b P_{\mathrm{block}}$ in full; describes the exact $F_2$-SCFG
inside--outside formulation that achieves it at $O(L^4)$;
re-encodes the same content as a memory-augmented Pair HMM at
$O(L^2 A^2)$; and develops the Gibbs+MH+replica-exchange MCMC
sampler that draws from the unbounded-edge limit at amortised
$O(L^3)$ per sweep. The conceptual hierarchy at the end of the
appendix lifts this construction to the full multi-branch
\emph{phylogenetic} SCFG, the natural fixed point of the entire
ladder of approximations.

\subsection{Exact 0-or-1-edge marginal posteriors via Pair-SCFG inside-outside}
\label{sec:f2-scfg}

The deprecated mean-field correction
(remark~\ref{rem:deprecated-mean-field} of
Appendix~\ref{app:tkfdp}) and the coupled-pair greedy annealer of
\S\ref{sec:coupled-annealing} are two routes to the same
intractable posterior of
equation~\ref{eq:tkfdp-joint-marginal}; the present subsection
derives an exact dynamic-programming alternative under the hard
restriction that the coupling partition $E$ contains zero or one
edge. The result is a per-residue-pair posterior $Q'_{ij}$ that
consumes the same downstream FSA assembler as the baseline, with
the coupling evidence absorbed exactly into the posterior rather
than fed in via a per-pair correction or via a custom annealer
move set.

\subsubsection*{Three pair-HMM marginal moments}

Under the no-coevolution baseline TKF92 Pair HMM with the
class-mixture single-site emissions of Appendix~\ref{app:tkfdp},
let $\pi(X, Y, A)$ denote the joint probability of an alignment
$A$ between sequences $X$ and $Y$. Define
\begin{align}
F_0(X, Y) \;&\equiv\; \sum_A \pi(X, Y, A), \label{eq:F0} \\
F_1(X, Y; i, j) \;&\equiv\; \sum_{A \,:\, X_i \sim_A Y_j} \pi(X, Y, A), \label{eq:F1} \\
F_2(X, Y; i, j; k, l) \;&\equiv\; \sum_{A \,:\, X_i \sim_A Y_j \,\wedge\, X_k \sim_A Y_l} \pi(X, Y, A), \label{eq:F2}
\end{align}
where $X_i \sim_A Y_j$ denotes that the alignment $A$ pairs residue
$X_i$ with $Y_j$ in a Match emission. Computationally, $F_0$ is a
Forward in $O(L^2)$, $F_1$ is a Forward--Backward in $O(L^2)$, and
$F_2$ is the Inside--Outside table of a non-bifurcating Pair SCFG
(or equivalently a doubled Forward--Backward) in $O(L^4)$ time and
memory. The conventional Pair-HMM match-state posterior is
$Q_{ij} \equiv F_1(i, j) / F_0$.

The objects $F_1$ and $F_2$ stand in the relation that gradients
and Hessians do to a partition function: $F_1(i, j) / F_0$ is the
marginal probability that $(X_i, Y_j)$ is paired and
$F_2(i, j; k, l) / F_0$ is the marginal joint probability that
both $(X_i, Y_j)$ and $(X_k, Y_l)$ are paired, with the
no-coevolution model serving as the unperturbed reference point.

\subsubsection*{The 0-or-1-edge Pair SCFG}

Let $M(i, j; k, l; t) \equiv M((X_i, X_k), (Y_j, Y_l); t)$ denote
the four-residue Potts coupling boost of
equation~\ref{eq:M-boost} of Appendix~\ref{app:tkfdp}, evaluated
at the observed residues $(X_i, Y_j, X_k, Y_l)$ and the inferred
branch length $t$. Define the (unnormalised) Pair-SCFG generating
function
\begin{equation}
\label{eq:scfg-grammar}
S \;\to\; S_0 \,\,\big|\,\, S_1, \qquad \text{with multiplicative weights $1$ and $\varepsilon = 1 / \alpha_z$,}
\end{equation}
where the two non-terminals expand as
\begin{align*}
S_0 \;&\to\; \text{TKF92 Pair HMM with zero coupled column-pairs (the baseline)}, \\
S_1 \;&\to\; \text{TKF92 Pair HMM with exactly one coupled column-pair}.
\end{align*}
The 1-edge nonterminal's parse picks one ordered column-pair
$(i, j; k, l)$ with $i < k$ as the coupled pair and replaces the
product of independent single-site Match emissions at columns
$(i, j)$ and $(k, l)$ with the joint pair-emission
$P_{\mathrm{joint}}((X_i, X_k) \to (Y_j, Y_l); t, H)$, equivalently
the product of independent emissions multiplied by
$M(i, j; k, l; t)$. The partition function of the grammar is
\begin{equation}
\label{eq:exact-01}
L_{\mathrm{exact}}(X, Y) \;=\; F_0(X, Y) \,+\, \varepsilon \sum_{(i, j; k, l), \,\, i < k} F_2(X, Y; i, j; k, l) \cdot M(i, j; k, l; t).
\end{equation}
The first term is the no-coupling marginal likelihood; the second
is the marginal contribution of all alignments that contain
exactly one coupled column-pair, summed over all candidate
locations $(i, j; k, l)$ for that pair. The $\varepsilon$ prefactor
is the per-pair partner prior under the size-$\{1, 2\}$ Ewens
partition with concentration $\alpha_z$; the assumption restricting
to at most one edge per alignment is a hard-truncation of the
size-$\{1, 2\}$ partition that retains the leading-order correction
in $\varepsilon$.

\subsubsection*{The exact match posterior}

Inside--Outside on the grammar~\eqref{eq:scfg-grammar} yields the
marginal probability that residue $X_i$ is aligned to $Y_j$ in the
data:
\begin{equation}
\label{eq:exact-Q}
Q'_{ij} \;=\; \frac{F_1(X, Y; i, j) \,+\, \varepsilon \sum_{(k, l) \,:\, k \neq i} F_2(X, Y; i, j; k, l) \cdot M(i, j; k, l; t)}{L_{\mathrm{exact}}(X, Y)}.
\end{equation}
The numerator collects all alignment-paths through the SCFG that
touch $(i, j)$ as a Match: the first term covers $S_0$ paths (no
coupled pair touches $(i, j)$), the second term covers $S_1$ paths
in which $(i, j)$ is part of the (unique) coupled pair, summed
over the choice of partner column $(k, l)$. Both summands include
the boost factor $M$ at exactly the right place. The denominator
is the partition function~\eqref{eq:exact-01}.

In the limit $\varepsilon \to 0$ (or $H = 0$ so $M \equiv 1$
everywhere), $Q'_{ij}$ collapses to $F_1(i, j) / F_0 = Q_{ij}$, the
no-coupling baseline, as required.

\subsubsection*{Cost and practical considerations}

The bottleneck of the explicit $F_2$ tensor encoding is the
$O(L^4)$ time and memory required to materialise $F_2$ itself:
$F_2$ has $\binom{L}{2}^2 \approx L^4 / 4$ free entries, which at
$L = 200$ occupies a few GB in single precision. Longer sequences
require either (a) prefix/suffix chunking on the second-anchor
axis $(k, l)$ to keep memory bounded, or (b) candidate pruning
that drops $(i, j)$ or $(k, l)$ pairs with $F_1(\cdot) < q_{\min}$,
reducing the candidate set from $O(L^4)$ to $O(L^2)$ in practice.
The reference implementation lives in \texttt{src/tkfdp/f2\_scfg.py};
in the current codebase it is retained primarily as a
cross-validation reference for the more efficient memory-augmented
Pair HMM described in the next subsection (which agrees with
$F_2$-SCFG to machine precision and runs at $O(L^2 A^2)$).

The available pathways to a coupled-FSA-consumable per-residue
posterior, in order of fidelity:

\begin{center}
\begin{tabular}{lll}
\textbf{Pathway} & \textbf{Cost} & \textbf{Approximation} \\
\hline
Mean-field pre-correction (deprecated; rem.~\ref{rem:deprecated-mean-field}) & $O(L^2 K_c^2 A^2)$ & First-order in $\varepsilon$ \\
Coupled-pair greedy annealer (\S\ref{sec:coupled-annealing}) & $O(L^3)$ & Greedy commit ordering \\
$F_2$-SCFG explicit enumeration & $O(L^4)$ & 0-or-1-edge truncation, otherwise exact \\
Aug-PHMM 1-edge (\S\ref{sec:aug-phmm}) & $O(L^2 A^2)$ & 0-or-1-edge truncation, otherwise exact \\
Aug-PHMM 2-edge (\S\ref{sec:aug-phmm}) & $O(L^2 A^4)$ & 0-or-1-or-2-edge truncation, otherwise exact \\
Infinite Pair HMM MCMC (\S\ref{sec:infinite-hmm}) & $O(L^4) + O(L^3) \cdot N_{\mathrm{sweep}}$ & Stochastic; converges to exact \\
\end{tabular}
\end{center}

The bounded-edge approximations capture progressively more of the
coupling correction at progressively higher cost; the infinite Pair
HMM MCMC is the principled limit, exact in expectation up to MCMC
sampling error. For families dominated by a single strong
coevolutionary edge per column, the 1-edge aug-PHMM at $O(L^2 A^2)$
is empirically adequate; for families with multiple competing
partners per column, the 2-edge aug-PHMM (or the MCMC sampler) is
necessary.

\subsection{Memory-augmented Pair HMM: the same content at $O(L^2 A^2)$}
\label{sec:aug-phmm}

The $F_2$-SCFG of equations~\ref{eq:scfg-grammar}--\ref{eq:exact-Q}
can be re-encoded as a Pair HMM whose latent state is augmented
with a small \emph{tag memory} of in-progress coupled-edge
endpoints. For the 0-or-1-edge case the tag set has
$|\mathcal{T}| = 1 + A^2 + 1$ values: a sentinel \texttt{no\_edge}
state, $A^2$ states $(a, b)$ recording that a left endpoint with
observed amino acids $(a, b)$ is awaiting its right partner, and a
terminal \texttt{done} state recording that a coupled pair has been
consumed. At any Match cell $(i, j)$ the HMM may either (a) carry
the tag forward unchanged, (b) transition
$\texttt{no\_edge} \to (X_i, Y_j)$ with weight $\varepsilon$ (spawn
a left endpoint), or (c) transition $(a, b) \to \texttt{done}$ with
weight $M(\cdot,\cdot;X_i,Y_j;t)$ summed over the recorded $(a, b)$
(close the pair). At all non-Match states the tag is preserved
verbatim. The augmented Forward partition function equals
$L_{\mathrm{exact}}$ of equation~\ref{eq:exact-01}, and the
marginal posterior at Match-state cell $(i, j)$, summed over tag
values, equals $Q'_{ij}$ of equation~\ref{eq:exact-Q}. Cost is
$O(L^2 \cdot 5 \cdot |\mathcal{T}|) = O(L^2 A^2)$ time and memory:
a factor $L^2 / A^2 \approx 100$ speedup over the explicit $F_2$
tensor at typical sequence lengths, with no loss of fidelity.
Implementation lives in \texttt{src/tkfdp/aug\_phmm.py};
cross-validation against the $O(L^4)$ $F_2$-SCFG reference
confirms agreement to machine precision (max
$|Q'_{ij} - Q'^{\mathrm{SCFG}}_{ij}| \approx 10^{-14}$ on tested
pairs).

The same construction generalises to allow up to $k$ coupled
column-pairs per alignment by enlarging the tag set to record
(i) the multiset of in-flight left endpoints (size up to $k$),
(ii) the number of resolved pairs (in $\{0, 1, \ldots, k\}$), and
(iii) optionally the in-flight endpoints alongside the resolved-pair
counter. For $k = 2$ the tag set has
$|\mathcal{T}_2| = 1 + A^2 + A^2(A^2+1)/2 + A^2 + 1 \approx 81{,}000$
values for $A = 20$: a sentinel \texttt{no\_edge}, $A^2$ singleton
states (one in-flight, none resolved), an unordered-multiset pair
state of size $A^2(A^2+1)/2$ (two in-flight, none resolved;
same-element multisets are valid distinct states), $A^2$
closed-singleton states (one resolved, one in-flight), and a
terminal \texttt{closed\_done}. The match-cell transitions add a
spawn from singleton$\to$pair with weight $\varepsilon$ (multiset
construction), and three closure transitions:
pair$\to$closed-singleton with weight $M(\cdot;X_i,Y_j;t)$ over the
closed multiset element, singleton$\to$closed-done with weight $M$,
and closed-singleton$\to$closed-done with weight $M$. Cost is
$O(L^2 \cdot 5 \cdot |\mathcal{T}_2|) = O(L^2 A^4)$ time and memory.

\subsection{The principled formulation: three-factor model and MCMC}
\label{sec:infinite-hmm}

The bounded-edge truncations (0-or-1-edge in \S\ref{sec:f2-scfg};
0-or-1-or-2-edge as a natural extension) are dynamic-programming
approximations to a more principled underlying model whose joint
posterior factorises cleanly into three independent ingredients:
the alignment path $A$ under the TKF92 Pair HMM, a partition
$\pi_{\!M}$ of its Match cells under a Chinese Restaurant Process
(CRP) prior, and the per-block substitution likelihood under the
substitution CTMC:
\begin{equation}
\label{eq:infinite-hmm-factorisation}
\pi(A, \pi_{\!M} \mid X, Y) \;\propto\; \pi_{\mathrm{TKF92}}(A \mid X, Y) \;\cdot\; \pi_{\mathrm{CRP}}(\pi_{\!M} \mid \alpha_z, |\mathrm{Match}(A)|) \;\cdot\; \prod_{b \,\in\, \pi_{\!M}} P_{\mathrm{block}}(b \mid t, H).
\end{equation}
Blocks of size~1 (uninvolved Match cells) contribute the standard
single-site CTMC factor $P_{\mathrm{singlet}}$ of
equation~\ref{eq:p-singlet} of Appendix~\ref{app:tkfdp}; blocks
of size~2 (coupled-edge endpoints) contribute the joint Potts
CTMC factor $P_{\mathrm{doublet}}$ of equation~\ref{eq:p-doublet}.
The CRP
prior with concentration $\alpha_z$ on a partition of
$N = |\mathrm{Match}(A)|$ Match cells into $K$ blocks of sizes
$\{n_b\}_{b=1}^K$ is the canonical Ewens partition
prior~\citep{ewens1972,pitmanyor1997}
\begin{equation}
\label{eq:crp-prior}
\pi_{\mathrm{CRP}}(\pi_{\!M} \mid \alpha_z, N) \;=\; \frac{\alpha_z^K \, \prod_{b=1}^K (n_b - 1)!}{\alpha_z (\alpha_z + 1) \cdots (\alpha_z + N - 1)};
\end{equation}
large $\alpha_z$ favours many small blocks (mostly singletons, few
coupled pairs). In practice we restrict to blocks of size at
most~2: blocks of size $\geq 3$ would correspond to triangular
cliques whose marginal mass is suppressed by factors of
$1/\alpha_z$ at high concentration. We call
\eqref{eq:infinite-hmm-factorisation} the \emph{infinite Pair HMM}
in the sense that the number of coupled edges per alignment is not
capped by the prior --- in contrast to the bounded-edge
dynamic-programming methods of \S\ref{sec:aug-phmm}.

The original \emph{infinite Hidden Markov Model} of Beal,
Ghahramani \& Rasmussen~\citep{bealghahramani2002} and the
HDP-HMM of Teh, Jordan, Beal \&
Blei~\citep{tehjordan2006} provide the broader Bayesian-nonparametric
framework in which equation~\ref{eq:infinite-hmm-factorisation}
sits.

\subsubsection*{MCMC sampler from the infinite Pair HMM}

The factorisation \eqref{eq:infinite-hmm-factorisation} admits an
efficient Markov-chain Monte Carlo sampler that scales as $O(L^4)$
for one-time setup followed by $O(L^3)$ amortised cost across many
sweeps. We give the setup phase first, then the three move types
that compose the sweep.

\paragraph{Target distribution.} The sampler operates on the joint
state $(A, E)$ where $A$ is a TKF92 Pair HMM alignment and $E$ is
a set of unordered Match-cell-pair edges. The unnormalised target
distribution is
\begin{equation}
\label{eq:mcmc-target}
\pi(A, E \mid X, Y) \;\propto\; P_{\mathrm{baseline}}(A \mid X, Y) \cdot \frac{\alpha_z^{|E|}}{\prod_{m=1}^{N_M(A)} (m - 1 + \alpha_z)} \cdot \prod_{e \in E} M(e \mid t),
\end{equation}
where $N_M(A)$ is the number of Match cells in $A$ and
$M(e \mid t)$ is the four-residue Potts boost at edge $e$. The
combinatorial prior is the canonical Ewens distribution restricted
to blocks of size $\leq 2$.

\paragraph{Setup phase.} The setup phase precomputes (a) the
partial-Forward tensor $F^{\mathrm{partial}}[i, j; k, l]$ giving
the baseline TKF92 Pair HMM probability of any alignment passing
through Match at both anchor cells $(i, j)$ and $(k, l)$ (the
same $O(L^4)$ object enumerated by the $F_2$ inside--outside
table of \S\ref{sec:f2-scfg}), and (b) the four-residue Potts
boost tensor $M$ at the observed amino acids over all anchor
quadruples. Both are stored once per $(X, Y)$ pair and reused
across all sweeps and chains.

\paragraph{The three MCMC moves.} The MCMC kernel composes three
move types:

\begin{enumerate}
\item \textbf{Segment-resample move (MH).} Pick a contiguous
  fragment of the current alignment $A$ bounded by two anchor
  Match cells $a_L < a_R$ (or by the alignment endpoints). Propose
  $A^{\mathrm{new}}$ by sampling the fragment from immediately
  after $a_L$ to immediately before $a_R$ by stochastic traceback
  on the segment-conditional Forward partition function
  $F^{\mathrm{partial}}_{\mathrm{segment}}[a_L; a_R]$ (Gibbs sample
  of the path conditional on the bounding anchors and the
  unchanged flanking alignment). Existing edges with endpoints in
  the resampled region are discarded; their orphaned partners
  outside the region become singletons in the partition. The
  proposal is exact under $\pi_{\mathrm{TKF92}}$, so the
  alignment-likelihood factor of \eqref{eq:mcmc-target} cancels in
  the Hastings ratio, leaving only the CRP and boost factors.

\item \textbf{CRP add-edge move (MH).} Pick a Match cell of $A$
  that is currently a singleton in $\pi_{\!M}$. Propose a partner
  Match cell from a proposal distribution
  $q_{\mathrm{add}}(\cdot \mid \text{singleton position})$ (the
  natural choice weights candidate partners by their boost
  magnitude $|M - 1|$). Insert the new edge into $E$. The Hastings
  ratio combines the CRP prior ratio
  $\alpha_z / (m - 1 + \alpha_z)$, the boost factor $M(e \mid t)$,
  and the reverse-proposal ratio (which is symmetric for a uniform
  remove-edge proposal of the same edge cardinality).

\item \textbf{CRP remove-edge move (MH).} Pick an existing edge in
  $E$ uniformly. Remove it; both endpoints become singletons. The
  Hastings ratio is the reciprocal of the add-edge ratio.
\end{enumerate}

\paragraph{Replica exchange.} Parallel-tempering replica exchange
over a ladder of $\alpha_z$ values (corresponding to varying
coupling strengths) ensures global mixing across the
multimodal posterior. Adjacent replicas swap states with the usual
Metropolis acceptance ratio for the temperature-pair exchange.

\paragraph{Cost.} The setup phase is $O(L^4)$ one-time per
sequence pair. Each MCMC sweep does $O(1)$ inter-edge-anchor
segment resamples; each segment resample is an $O(L_{\rm seg})$
stochastic traceback through the cached partial-Forward tensor,
giving an amortised $O(L)$ per-sweep cost. Aggregated
per-cell match indicators across post-burn-in samples yield a
stochastic estimate of the per-residue posterior $Q'_{ij}$ that
the FSA assembler consumes in place of the bounded-edge analytic
estimate. The block move reduces to a pure Gibbs ``segment
between adjacent edge anchors'' resample whenever $a_L$ and $a_R$
are adjacent in the current edge-anchor list.

\subsubsection*{Algorithmic detail}

\begin{algorithm}[H]
\caption{One MCMC sweep on the infinite Pair HMM}
\label{alg:mcmc-sweep}
\begin{algorithmic}[1]
\Require Current state $(A, E)$, partial-Forward tensor
  $F^{\mathrm{partial}}$, boost tensor $M$, concentration
  $\alpha_z$
\Ensure New state $(A', E')$
\State Pick $K_{\mathrm{seg}}$ segment-resample positions
  (e.g.\ $K_{\mathrm{seg}} = 8$, distributed by alignment length)
\For{each segment $[a_L, a_R]$}
  \State $A_{\mathrm{seg}}^{\mathrm{new}} \gets$ stochastic
    traceback of $F^{\mathrm{partial}}_{\mathrm{segment}}[a_L; a_R]$
  \State Discard edges in $E$ with at least one endpoint in
    $[a_L, a_R]$ (orphaned partners become singletons)
  \State $\rho \gets$ MH ratio combining CRP and boost factors
    (alignment factor cancels by construction of the proposal)
  \If{$u \sim \mathrm{Uniform}(0, 1) < \rho$}
    \State $A \gets A$ with $[a_L, a_R]$ replaced by
      $A_{\mathrm{seg}}^{\mathrm{new}}$
    \State $E \gets E$ with orphaned edges removed
  \EndIf
\EndFor
\State Pick $K_{\mathrm{add}}$ add-edge candidates and
  $K_{\mathrm{rem}}$ remove-edge candidates (e.g.\
  $K_{\mathrm{add}} = 4 |E|$, $K_{\mathrm{rem}} = |E|$)
\For{each candidate}
  \State Propose under the appropriate move kernel
  \State Compute MH ratio (CRP $\times$ boost for add;
    reciprocal for remove)
  \State Accept / reject
\EndFor
\State Periodically: attempt replica-exchange swap with adjacent
  $\alpha_z$ rung
\State \Return $(A, E)$
\end{algorithmic}
\end{algorithm}

\noindent The exact form of the segment-resample MH ratio depends
on the edge-handling convention. Under the simplest convention
(``Strategy~S-1'' in the implementation spec), edges whose
endpoints lie wholly within the resampled region are discarded
unconditionally and their boost contributions enter only the
forward MH ratio; edges with one endpoint outside the region are
discarded and their orphaned partner becomes a singleton.

\subsubsection*{Verification protocol}
\label{sec:phmm-mcmc-verification}

The MCMC sampler is verified against four reference points:

\begin{description}
\item[E.1 Cross-validate against \texttt{aug\_phmm} (1-edge, large
  $\alpha_z$).] At large $\alpha_z$ the infinite Pair HMM truncates
  to a 0-or-1-edge model, and the per-cell match-marginal
  $Q'_{ij}$ produced by the MCMC sampler should agree with the
  exact $O(L^2 A^2)$ aug-PHMM 1-edge inside--outside.
\item[E.2 Cross-validate against \texttt{aug\_phmm\_2edge} (2-edge,
  moderate $\alpha_z$).] At moderate $\alpha_z$ the truncation
  extends to 0-or-1-or-2-edge; agreement with the exact
  $O(L^2 A^4)$ aug-PHMM 2-edge inside--outside is the next
  consistency check.
\item[E.3 Cross-validate against brute-force enumeration.] On
  small sequence pairs $(L_x, L_y) \leq (5, 4)$ where all
  alignments and all $\leq 5$-edge placements can be enumerated
  directly, the MCMC sample mean of $Q'_{ij}$ must converge to the
  brute-force value at the canonical $1/\sqrt{n}$ rate.
\item[E.4 Detailed-balance sanity check (no-data case).] At
  $H = 0$ (so $M \equiv 1$) the boost factor is identically one,
  the per-block factor reduces to the baseline emission, and the
  sampler should recover the no-edge TKF92 Pair HMM marginal
  exactly --- with edge-set sample sizes distributed as the
  prior Ewens distribution.
\end{description}

The first three checks pin the sampler to the exact reference at
each of three orders of edge truncation; the fourth checks
detailed balance independently of the data-conditioned target.

\subsection{The conceptual hierarchy: infinite phylogenetic SCFG}
\label{sec:phylo-scfg}

\paragraph{The infinite Pair HMM is itself an approximation.} The
infinite Pair HMM of equation~\ref{eq:infinite-hmm-factorisation}
is the principled limit of the bounded-edge dynamic-programming
approximations of \S\ref{sec:aug-phmm}, but it is itself an
approximation to a still-more-principled object: the
infinite-edge generalisation of the gravestone-augmented
time-indexed pair SCFG of \S\ref{sec:gravestone-scfg} of
Appendix~\ref{app:tkfdp}. That fuller model samples not only the
visible alignment $A$ and the edge set $E$ but also the timed
gravestone-augmented branch history $G$ comprising every fragment
that ever existed during the branch (whether visible at the
alignment endpoints, deleted on the branch, or transiently
inserted then deleted), with the same canonical CRP rule of
equation~\ref{eq:infinite-hmm-factorisation} governing edge spawn
at every Match-class member of either an alive or a gravestone
fragment. On the substitution side, transient gravestones inserted
during the branch can be assigned the same $z_s$ key-DP class as
their alive class-mates and therefore contribute coupling terms to
those class-mates' rates during the gravestone's lifetime --- so
the substitution likelihood at a Match cell $(i, j)$ depends on
$G$, not just on $A$. The infinite Pair HMM of this appendix
marginalises $G$ implicitly via the standard TKF92 Pair HMM
substitution emissions, which is correct in expectation when
transient lineages are rare or when their coupling contribution is
small, but loses the within-branch dynamics that the
gravestone-augmented SCFG captures.

The resulting hierarchy is nested: the bounded-edge
dynamic-programming methods of \S\ref{sec:aug-phmm} are exact in
the limit of the infinite Pair HMM of this section; the infinite
Pair HMM is exact in the limit of the infinite gravestone-augmented
pair SCFG; and that fuller object is the natural fixed point at
which the indel-side and the substitution-side principled
formulations meet. The MCMC sampler described above samples from
the middle level of this hierarchy; lifting it to the full
infinite gravestone-augmented pair SCFG --- by interleaving the
infinite-HMM CRP edge moves with the gravestone-augmented
branch-history moves of \S\ref{sec:gravestone-scfg} of
Appendix~\ref{app:tkfdp} --- is a natural extension.

\paragraph{Further: the infinite gravestone-augmented
\emph{phylogenetic} SCFG.} The pairwise hierarchy above lifts to
the full multi-branch phylogenetic setting at no further modelling
cost, by composition with the Felsenstein-style upwards--downwards
machinery of \S\ref{sec:gravestone-scfg} that already glues
per-branch gravestone-augmented histories into a tree-level
posterior. At every internal node and along every branch of the
tree, the gravestone-augmented latent state is sampled jointly
with the alignment, while the same canonical CRP rule of
equation~\ref{eq:infinite-hmm-factorisation} governs edge spawn at
every Match-class member of every fragment (alive or gravestone)
at every node. The MCMC sampler above lifts by interleaving (i)
per-branch gravestone-augmented history moves of
\S\ref{sec:gravestone-scfg}, (ii) per-class-pair CRP edge moves of
this appendix, and (iii) the per-internal-node Felsenstein-style
updates already present in the SVI loop of
\S\ref{sec:gravestone-scfg}. Each ingredient is already
load-bearing in the existing pipeline; no new modelling assumptions
are required to assemble them.

The full conceptual ladder is therefore:
\begin{quote}
mean-field pre-correction (deprecated; rem.~\ref{rem:deprecated-mean-field}) $\to$
bounded-edge aug-PHMM (\S\ref{sec:aug-phmm}) $\to$
infinite Pair HMM (\S\ref{sec:infinite-hmm}) $\to$
infinite gravestone-augmented pair SCFG
  (\S\ref{sec:infinite-hmm} $+$ \S\ref{sec:gravestone-scfg}) $\to$
infinite gravestone-augmented \emph{phylogenetic} SCFG
  (\S\ref{sec:phylo-scfg}),
\end{quote}
each rung exact in the limit of the next, the entire ladder
falling out of the simple stick-breaking-TKF92-with-Potts
construction of Appendices~\ref{app:tkfdp}--\ref{app:phmm} with no
further mathematical machinery.
