% =============================================================
% Appendix E: TKF-DP --- Dirichlet-process Potts coupling
% Source: ~/tkf-dp/main.tex Sections 2 and 3, quoted near-verbatim.
% Section numbering and labels have been adapted to fit the
% supplement's overall scheme. References to the present
% appendix's own equations are local; references to other
% appendices use their supplement-level labels.
% =============================================================

The TKF-DP model decorates TKF92 with three independent Dirichlet
processes that govern (i) the per-site key partition into Potts
coevolutionary cliques; (ii) the per-site assignment to a
substitution class; and (iii) the per-class-pair assignment to a
Potts coupling atom. The class-level substitution dynamics is a
multi-site reversible CTMC; the alignment likelihood remains plain
TKF92. Inference proceeds by a variational E-step on the
continuous-time class clique, Gibbs sweeps and Jain--Neal
split--merge proposals on the three Dirichlet partitions, recursive
traceback over a time-indexed gravestone-augmented pair SCFG for
the augmented branch history, and a stochastic M-step on the
corpus-level hyperparameters with closed-form
Dirichlet--multinomial / Gamma--Poisson / Laplace conjugacies
unlocking the new-class / new-atom base-measure integrals.

\subsection{The TKF-DP generative model}
\label{sec:tkfdp-model}

The TKF92 model~\citep{ThorneEtAl92} of insertion, deletion, and
substitution has the convenient property that the alignment path
and the substitution history factorize: the marginal distribution
over alignments is independent of the amino acid dynamics, and
inference proceeds via a pair HMM on the alignment with substitution
likelihoods plugged in pointwise. This factorization is exactly
what is lost when site--site coevolutionary couplings are introduced
naively, since the substitution process at any site then depends on
the joint state of its coupled partners, and any coupling rule that
consults the existing alignment to decide partners destroys the
marginal pair HMM.

We describe a minimal extension of TKF92 that re-introduces
site--site Potts couplings while preserving the
alignment--substitution factorization. Each newborn site is assigned
a discrete key drawn i.i.d.\ from a Dirichlet process, and sites
that share a key co-evolve under a Potts-modulated joint CTMC.
Because the keys are drawn independently of the alignment, the
alignment marginal is unchanged.

The alignment--substitution factorization survives at the parameter
level but, once couplings are present, breaks at the latent-variable
level: the substitution likelihood conditional on the alignment
depends on the full timed indel history, including transient
(alignment-invisible) lineages whose presence transiently modifies
the rates of co-class members. Inference therefore requires a
tractable representation of timed indel histories. We obtain one
via a gravestone-augmented latent state and a time-indexed pair
SCFG whose inside--outside is closed form
(\S\ref{sec:gravestone-scfg}).

\begin{definition}[TKF-DP]
\label{defn:tkfdp}
Fix TKF92 indel parameters $(\lambda, \mu, r)$, a key concentration
parameter $\alpha_z > 0$, and a symmetric $20 \times 20$ Potts
coupling matrix $H$. The process jointly generates a TKF92
alignment path, a key assignment for every site, and an amino acid
trajectory at every site, as follows.
\begin{enumerate}
    \item Sample stick-breaking weights once, globally:
    \[
        \pi_k = \beta_k \prod_{l < k} (1 - \beta_l),
        \qquad
        \beta_k \stackrel{\text{iid}}{\sim} \mathrm{Beta}(1, \alpha_z).
    \]
    \item Run TKF92 indel dynamics with parameters $(\lambda, \mu, r)$,
      giving a stochastic set of alive sites $S(t)$ over time.
    \item At the birth of each new site $s$, draw a key
    \[
        z_s \mid \pi \sim \mathrm{Cat}(\pi),
    \]
    independently of all other sites, of the alignment path, and of
    the substitution history. The key is fixed for the site's
    lifetime.
    \item At any time $t$, the alive sites $S(t)$ are partitioned
      into equivalence classes $\{C_k(t)\}$ by the relation $z_s =
      z_{s'}$. The amino acid states $\{x_s\}_{s \in S(t)}$ evolve
      as independent CTMCs across classes; the CTMC on class $C$ is
      a $|C|$-site reversible chain with stationary distribution
    \[
        \pi_C(x_C) \propto \exp\!\Big(-\!\!\sum_{\{i,j\} \subset C} H(x_i, x_j)\Big).
    \]
    Class composition is piecewise constant between birth/death
    events; on each event, the affected class is rebuilt and the
    surviving sites re-equilibrate under the new class generator.
\end{enumerate}
\end{definition}

\begin{remark}[Factorization]
Conditional on $\pi$, the key draws are i.i.d.\ across sites, so the
joint distribution factorizes as
\[
    P(\mathrm{alignment},\, \pi,\, \{z_s\},\, \mathrm{substitution})
    = P(\mathrm{alignment}) \cdot P(\pi) \cdot \prod_s P(z_s \mid \pi)
      \cdot P(\mathrm{substitution} \mid \mathrm{alignment}, \{z_s\}).
\]
In particular, the marginal alignment likelihood is exactly that of
TKF92, and the pair HMM one would use under $\alpha_z = \infty$ (no
coupling) is unchanged.
\end{remark}

\begin{remark}[Partition statistics]
Marginalizing $\pi$, the partition induced on any fixed finite set
of sites is the Ewens partition with parameter $\alpha_z$,
equivalent to a Chinese restaurant process draw~\citep{aldous1985}.
The expected number of co-evolving classes among $L$ sites is
$\mathcal{O}(\alpha_z \log L)$, and tuning $\alpha_z$ controls the
sparsity of couplings: $\alpha_z \gg L$ yields
$\mathcal{O}(L^2 / \alpha_z)$ co-evolving pairs with higher-order
classes suppressed by additional factors of $L / \alpha_z$. Smaller
$\alpha_z$ gives Ewens-distributed class sizes with heavier tails,
including occasional larger cliques.
\end{remark}

\subsection{IBP variant}
\label{sec:tkfdp-ibp}

To allow each site to participate in multiple independent couplings,
replace the Dirichlet process with an Indian Buffet
Process~\citep{ibp}.

\begin{definition}[TKF92-IBP]
Sample feature frequencies $\{\pi_k\}_{k \geq 1}$ from an IBP with
concentration $\alpha_z$. At the birth of each site $s$, draw a
binary feature vector $b_s \in \{0, 1\}^{\mathbb{N}}$ with
\[
    b_{s,k} \mid \pi_k \stackrel{\text{ind}}{\sim} \mathrm{Bern}(\pi_k),
\]
independently across sites. Two alive sites $s, s'$ are coupled iff
there exists $k$ with $b_{s,k} = b_{s',k} = 1$. The amino acid
dynamics is the reversible CTMC on the alive sites with stationary
distribution
\[
    \pi(x_S) \propto \exp\!\Big(-\!\!\sum_{\{s, s'\} \,:\, \exists k,\, b_{s,k} = b_{s',k} = 1} H(x_s, x_{s'})\Big).
\]
\end{definition}

The same factorization observation applies: feature draws are i.i.d.\
given $\{\pi_k\}$, so the marginal alignment likelihood remains
plain TKF92. The IBP variant naturally accommodates hub structure
(residues that interact with many partners) via heavy-tailed feature
frequencies.

\subsection{Site classes and a GTR-parameterized generator}
\label{sec:gtr-potts}

Real proteins exhibit substantial heterogeneity in equilibrium amino
acid usage and in coupling profile: positions in $\alpha$-helices,
$\beta$-sheets, buried cores, and disordered regions have distinct
frequencies and distinct interaction patterns with their structural
neighbours. We extend the model with a per-site latent class
variable that determines both the uncoupled (point) substitution
model and the per-pair Potts interaction tensor. The variable is
non-evolving, drawn once at site birth, and we put a second
Dirichlet process on it --- independent of the key DP --- so that
the number of site classes is itself learned from data, in line
with the original infinite-mixture CAT model of Lartillot and
Philippe~\citep{lartillotcat}.

\begin{definition}[Site-class TKF-DP]
Fix the alphabet $\mathcal{A}$ with $|\mathcal{A}| = A = 20$, a
symmetric exchangeability matrix
$S \in \mathbb{R}^{A \times A}_{\geq 0}$ ($S_{xy} = S_{yx}$),
per-site rate-multiplier hyperparameters $(a_\eta, b_\eta)$, a
site-class concentration $\alpha_c > 0$, a Potts-atom concentration
$\alpha_H > 0$, and base measures
\[
    G_0: \quad \pi^{(c)} \stackrel{\text{iid}}{\sim}
              \mathrm{Dirichlet}(\kappa_\pi\, \bar\pi),
    \qquad
    G_0^H: \quad H_h(i, j) \stackrel{\text{ind}}{\sim}
              \mathcal{N}\!\bigl(\mu_{kl},\, \tau_{kl}^{-1}\bigr)
              \;\text{for } i \leq j,\;
              H_h(j, i) := H_h(i, j),
\]
where $(k, l) = (\min(i, j), \max(i, j))$ index the $A(A+1)/2$
unordered amino-acid pairs, and each Potts atom
$H_h \in \mathbb{R}^{A \times A}$ is symmetric in $(i, j)$ by
construction. Site classes are drawn from a Dirichlet process
$D_c \sim \mathrm{DP}(\alpha_c, G_0)$ via stick-breaking; Potts
atoms are drawn from a separate Dirichlet process
$D_H \sim \mathrm{DP}(\alpha_H, G_0^H)$. At the birth of each site
$s$, draw an independent per-site rate multiplier and a site class:
\[
    \eta_s \stackrel{\text{iid}}{\sim} \mathrm{Gamma}(a_\eta, b_\eta),
    \qquad
    c_s \mid D_c \stackrel{\text{iid}}{\sim} D_c,
\]
in addition to the DP key $z_s$
(\S\ref{sec:tkfdp-model}); $\eta_s$, $c_s$, and $z_s$ are mutually
independent and independent of the alignment path, and are fixed
for the site's lifetime. The Potts coupling tensor is decomposed by
class-pair: for each unordered pair of classes $\{c, c'\}$ that ever
co-occur within a key DP class, draw a Potts-atom assignment
\[
    h_{cc'} \mid D_H \stackrel{\text{iid}}{\sim} D_H,
\]
at first co-occurrence and fixed thereafter; the materialized slice
for the pair is $H_{c c'}(i, j) := H_{h_{cc'}}(i, j)$. The
amino-acid trajectory on a DP key class $C$ is the single-site
reversible CTMC with stationary distribution
\[
    \pi_C(x_C) \;\propto\; \prod_{s \in C} \pi^{(c_s)}(x_s) \cdot
    \exp\!\Big(-\!\!\sum_{\{s, s'\} \subset C}\!\! H_{c_s c_{s'}}(x_s, x_{s'}) \Big),
\]
and (taking the F81 instance reversible with respect to this
stationary)
\[
    Q^s\!\bigl(x \to x'; x_{C \setminus s}\bigr)
    \;=\; \eta_s\, S_{xx'}\, \pi^{(c_s)}(x')\,
    \exp\!\Bigl(-\tfrac{1}{2} \Delta H_s\Bigr),
    \qquad
    \Delta H_s = \!\!\sum_{s' \in C \setminus s}\!\![H_{c_s c_{s'}}(x', x_{s'}) - H_{c_s c_{s'}}(x, x_{s'})].
\]
\end{definition}

\begin{remark}[GTR limit and singletons]
For a singleton key DP class, $\Delta H_s \equiv 0$ and the rate
reduces to $Q^s_{xy} = \eta_s S_{xy} \pi^{(c_s)}(y)$, the F81
instance of GTR with class-specific equilibrium and a per-site rate
multiplier. Marginalized over $c_s$ this is exactly the CAT
infinite-mixture profile model of Lartillot and
Philippe~\citep{lartillotcat}, augmented with Yang's gamma per-site
rate variation~\citep{Yang1994a}. Coupling enters only via
$\Delta H_s$ when the key DP class size exceeds one, so the model
strictly extends the CAT--gamma combination by adding a
coevolutionary tail.
\end{remark}

\begin{remark}[F81 vs.\ symmetric-Metropolis form]
The F81 form $Q_{xy} \propto S_{xy} \pi(y)$ chosen above is one of
two natural reversible instances of the GTR family with stationary
$\pi$ and exchangeability $S$; the other is the symmetric-Metropolis
instance $Q_{xy} \propto S_{xy} \sqrt{\pi(y)/\pi(x)}$. Both satisfy
detailed balance with respect to the joint Potts stationary above.
The symmetric-Metropolis form has the property that the
eigendecomposition of its symmetrized similarity transform is
independent of $\pi$, allowing one corpus-wide $A^3$ decomposition
to be reused across all classes. We adopt the F81 form here because,
combined with the secret-destination augmentation of
\S\ref{sec:base-measure-conj}, it yields strict
Dirichlet--multinomial conjugacy on $\pi^{(c)}$, which the
symmetric-Metropolis form does not. The eigendecomposition then
becomes per-class (one $A^3$ per class per outer step), which is
small at $A = 20$ and is dominated by the Potts cluster work in any
case.
\end{remark}

\begin{remark}[Three independent Dirichlet processes]
The construction has \emph{three} Dirichlet processes acting on the
substitution side: the key DP with concentration $\alpha_z$
partitions sites into co-evolving cliques (\S\ref{sec:tkfdp-model});
the site-class DP with concentration $\alpha_c$ assigns each site a
profile $\pi^{(c)}$; and a Potts-atom DP with concentration
$\alpha_H$ and base measure $G_0^H$ assigns each unordered class-pair
$\{c, c'\}$ to a coupling atom $H_{h_{cc'}}$. With $\alpha_H$ small
the $K(K+1)/2$ class-pair indices collapse onto a small number of
canonical Potts atoms shared across class-pairs; with
$\alpha_H \to \infty$ each class-pair has its own atom. The
Potts-atom DP is what makes the model identifiable for $K_c \gg 1$:
without it, the number of free Potts parameters scales as
$K_c^2 \cdot A^2$; with it the count scales as
$K_H \cdot A^2 + K_c^2$ class-pair indices, where $K_H$ is
data-driven and typically much smaller than $K_c^2$.
\end{remark}

\begin{remark}[Per-class-pair side potentials]
\label{rem:side-potentials}
A useful refinement of the per-class profile $\pi^{(c)}$ is to give
each unordered class-pair $\{c, c'\}$ its own pair of \emph{side
potentials}
$h^{(c, c')}_a, h^{(c, c')}_b \in \mathbb{R}^A$, with independent
Gaussian priors centered at zero,
\begin{equation}
h^{(c, c')}_a(x), h^{(c, c')}_b(y) \;\sim\; \mathcal{N}(0, \tau_h^{-1}).
\end{equation}
The per-pair joint stationary then becomes
\begin{equation}
\log \pi^{(c, c')}_{\mathrm{joint}}(x, y) \;=\; \log \pi^{(c)}(x) + \log \pi^{(c')}(y) - h^{(c, c')}_a(x) - h^{(c, c')}_b(y) - H_{cc'}(x, y) - \log Z_{cc'},
\end{equation}
i.e.\ $h$ is the per-pair deviation of the effective per-site
stationary
$\pi^{\mathrm{eff}, a}_{cc'}(x) \propto \pi^{(c)}(x) e^{-h^{(c, c')}_a(x)}$
from the per-class background $\pi^{(c)}$. For self-pairs $(c, c)$
the two sites are exchangeable, so the joint distribution must be
symmetric under swap $(x, y) \leftrightarrow (y, x)$. Combined with
$H_{cc}$ symmetric in its arguments, this forces
$h^{(c,c)}_a = h^{(c,c)}_b$; we parameterise self-pairs with a
single shared vector $h^{(c,c)} \in \mathbb{R}^A$ (and prior
$h^{(c,c)} \sim \mathcal{N}(0, \tau_h^{-1})$, contributing $A$ free
parameters per self-pair rather than $2A$). Off-diagonal pairs
$\{c, c'\}$ with $c \ne c'$ retain two independent vectors
$h^{(c,c')}_a, h^{(c,c')}_b$ ($2A$ free parameters each).
Reversibility of the F81-form joint generator is preserved by
replacing the destination factor $\pi^{(c)}(x')$ in the site-1 flip
rate with $\pi^{\mathrm{eff}, a}_{cc'}(x')$ (and analogously for
site 2).

The motivation is that $\pi^{(c)}$ captures the population-mean
amino-acid usage of class $c$ \emph{averaged over all class-pair
contexts}, while the actual usage of class $c$ when paired with
class $c'$ may deviate (e.g.\ a Cys-bearing class paired with itself
is more Cys-rich than the per-class average, because we are
conditioning on disulfide-forming sites). The $h$ vectors absorb
that deviation, leaving the Potts atom $H_{cc'}$ to capture the
coupling \emph{above the marginal compositional shift}. The Gaussian
prior pulls $h$ to zero so that any pair-specific deviation must be
data-supported.
\end{remark}

\begin{remark}[Hierarchical structure of the $H$ base measure]
The amino-acid-pair-indexed Gaussian base measure
$\mathcal{N}(\mu_{kl}, \tau_{kl}^{-1})$ embeds an empirical-Bayes
hierarchy: the corpus-level $(\mu_{kl}, \tau_{kl})$ encode the
population-mean coupling propensity for each amino-acid pair
(typically attractive for hydrophobic--hydrophobic,
charge-complementary, etc., and repulsive for charge-like-like), and
atom-specific deviations $H_h(i, j) - \mu_{kl}$ are learned around
that mean with atom-specific magnitude controlled by the data.
Symmetry of each $H_h$ in $(i, j)$ encodes the unordered-edge
convention of Potts couplings; this is standard practice in the DCA
literature~\citep{ekeberg2013} and is biologically natural in the
absence of any a priori labelling distinguishing the two ends of an
inter-class edge. The Gaussian prior on each $H_h$ is conjugate to
the Gaussian (Hessian-based) Laplace approximation of the path-DCA
likelihood, giving a closed-form Gaussian posterior on each
materialized atom under the local quadratic expansion of
\S\ref{sec:base-measure-conj}.
\end{remark}

\subsection{Class-level variational substitution likelihood}
\label{sec:class-elbo}

Inference under the model requires, for each branch of a phylogeny
and each co-evolving DP class $C$, the per-branch transition
probability $P(\mathbf{x}_C(t) \mid \mathbf{x}_C(0), \{c_s\}_{s \in C})$.
Direct evaluation requires exponentiating a generator on
$\mathcal{A}^{|C|}$ and is intractable beyond $|C| \approx 5$. The
bulk of the key-DP partition is small-class for $\alpha_z$ in the
practical regime, so exact computation handles most of the work;
what remains is to control the tail of large classes.

We derive a variational lower bound from an exact path-measure
factorization on the class clique. The factorization is the analog,
on path space, of the static pseudolikelihood approximation used in
DCA inference~\citep{ekeberg2013}, but with the static partition
function entirely absorbed into the conditioned-on initial state.
Two variational families on the resulting Girsanov ELBO are natural.
The first is the simpler scheme we develop here, parameterized by
per-site distributions at internal tree nodes plus a single constant
rate matrix per site per branch, with a closed-form ELBO via
eigendecomposition and a Felsenstein-style pruning recursion for
optimization. The second is the fully inhomogeneous mean-field
scheme of Cohn et al.~\citep{cohn2010}, with time-varying rates and
forward-backward ODE sweeps, which we describe as an upgrade for
branches where the constant-rate scheme is too loose.

\begin{proposition}[Path-measure factorization]
Let $\sigma_C \in D([0,t]; \mathcal{A}^{|C|})$ be the trajectory of
a single-site reversible CTMC on class $C$ with rate
$Q^s(x_s \to x_s'; x_{C \setminus s})$ and stationary distribution
$\pi_C$. The path measure satisfies
\[
    \mathbb{P}(\sigma_C) \;=\;
    \pi_C\!\bigl(\sigma_C(0)\bigr)
    \prod_{s \in C} \mathbb{P}_s\!\bigl(\sigma_s \,\big|\, \sigma_{C \setminus s}\bigr),
\]
where $\mathbb{P}_s(\cdot \mid \sigma_{C \setminus s})$ is the law
of an inhomogeneous CTMC on site $s$ with rate
$Q^s\!\bigl(\,\cdot\, ; \, x_{C \setminus s}(u)\bigr)$ at time $u$.
\end{proposition}

\noindent The factorization is exact: simultaneous jumps at distinct
sites have zero Lebesgue measure, every jump is therefore
attributable to a unique site, and the path log-density splits
site-wise. The only intractable factor is $\pi_C(\sigma_C(0))$, which
is conditioned upon for transition probabilities and so cancels.

\subsubsection*{Constant-rate variational family}

\begin{remark}[Variational family]
The variational family is parameterized at two levels. At each
internal tree node $v$ and site $s$, a discrete distribution
$\nu_v^s \in \Delta^{|\mathcal{A}|-1}$ over residue states; node
distributions factorize over sites,
$\nu_v(\mathbf{x}_v) = \prod_s \nu_v^s(x_v^s)$. At each branch $b$
between nodes $p$ and $c$ of length $t_b$ and each site $s \in C$,
a constant rate matrix $\hat Q^s_b$ on $[0, t_b]$. Conditional on
sampled endpoint states $(x_p^s, x_c^s) \sim \nu_p^s \otimes \nu_c^s$,
the variational law on the branch is the CTMC bridge of
$\hat Q^s_b$ from $x_p^s$ to $x_c^s$, with sites independent given
endpoints. The joint variational law on the tree is
\[
    \mathbb{Q} = \prod_v \nu_v \cdot \prod_b \prod_{s \in C}
    \mathbb{Q}^s_b\bigl(\sigma^s_b \,\big|\, x_p^s, x_c^s\bigr).
\]
\end{remark}

\begin{proposition}[Closed-form per-branch ELBO]
The path-measure KL contribution from branch $b$ and site $s$ has
Girsanov form
\[
    \mathrm{KL}_{b, s} = \mathbb{E}_{\mathbb{Q}}\!\!\int_0^{t_b}\!\!
    \sum_{x, x'} q^s_b(x; u)
    \!\Bigl[\, \mathbb{E}_{u}\bigl[Q^s(x \to x'; x_{C \setminus s}(u))\bigr]
    - \hat Q^s_b(x \to x')
    + \hat Q^s_b(x \to x') \log\!\frac{\hat Q^s_b(x \to x')}{\exp \mathbb{E}_{u}[\log Q^s(\cdot)]} \Bigr]\, du,
\]
where $q^s_b(x; u) = \mathbb{E}_{x_p, x_c}[q^s_b(x; u \mid x_p, x_c)]$
is the variational bridge marginal of site $s$ at time $u$ averaged
over endpoint distributions, and the inner expectations
$\mathbb{E}_{u}[\cdot]$ are over the product of \emph{instantaneous}
neighbour bridge marginals at $u$,
$\prod_{s' \neq s} q^{s'}_b(\cdot; u)$. The integrand is therefore a
product of two time-varying bridge marginals at the same instant
$u$, and the integral over $[0, t_b]$ does \emph{not} factorise into
time-averages: the cross-temporal correlation between
$q^s_b(\cdot; u)$ and $q^{s'}_b(\cdot; u)$ contributes to
$\mathrm{KL}_{b,s}$ and is what makes the formula a strict lower
bound. The bridge-expectation closed form below evaluates the integral
correctly.

In the constant-rate family, the stationary point of
$\mathrm{KL}_{b,s}$ in $\hat Q^s_b$ admits no closed form in general
because of the cross-temporal correlation. The geometric-mean rate
\[
    \log \hat Q^{s, *}_b(x \to x') \;=\;
    \mathbb{E}_{\bar q^{C \setminus s}_b}\!\bigl[\log Q^s(x \to x'; x_{C \setminus s})\bigr],
\]
where $\bar q^{s'}_b(x) = (1/t_b) \int_0^{t_b} q^{s'}_b(x; u) du$
is the time-averaged neighbour distribution for each $s' \neq s$,
is the stationary point under the additional ``mean-field-of-bridges''
assumption that $q^s_b(\cdot; u)$ and $q^{s'}_b(\cdot; u)$ are
approximately uncorrelated in $u$ on $[0, t_b]$. This assumption is
exact at $H = 0$ (sites independent under the variational law as
well as under the truth), close to exact for small coupling, and
only mildly violated at typical phylogenetic time scales; we take
$\hat Q^{s, *}_b$ above as the chosen variational rate and evaluate
the strict $\mathrm{KL}_{b, s}$ at this rate via the bridge-expectation
formula in the Remark below. Each term of the resulting bound is
$\geq 0$ by Jensen, so $\mathrm{KL}_{b, s} \geq 0$ and the ELBO is a
strict lower bound on $\log P$, as required.
\end{proposition}

\begin{remark}[Closed-form ELBO via pairwise bridge expectations]
\label{rem:bridge-expectations}
Each rate $\hat Q^s_b$ admits an eigendecomposition
$\hat Q^s_b = U^s D^s (V^s)$ with $V^s = (U^s)^{-1}$ and eigenvalues
$\{\lambda^s_k\}$ (we drop the branch subscript $b$ for legibility).
The forward and backward propagators have the same expansion:
\[
    [\exp(\hat Q^s u)]_{x_p, x} = \sum_k U^s_{x_p, k}\, V^s_{k, x}\, e^{\lambda^s_k u},
    \quad
    [\exp(\hat Q^s (t-u))]_{x, x_c} = \sum_l U^s_{x, l}\, V^s_{l, x_c}\, e^{\lambda^s_l (t-u)},
\]
so the per-site bridge marginal is
\[
    q^s_b(x; u \mid x_p, x_c) =
    \frac{1}{P_s} \sum_{k, l} U^s_{x_p, k}\, V^s_{k, x}\,
    U^s_{x, l}\, V^s_{l, x_c}\, e^{\lambda^s_k u}\, e^{\lambda^s_l (t-u)},
\]
with $P_s = [\exp(\hat Q^s t)]_{x_p, x_c}$ the bridge normalising
constant. The branch integrand of $\mathrm{KL}_{b, s}$ becomes a sum
over $(k, l)$ and over $(k', l')$ from the neighbour expansion, with
pair-time integrals
\[
    \mathrm{HR}(\alpha, \beta, t) \;=\;
    \begin{cases}
    (e^{\alpha t} - e^{\beta t}) / (\alpha - \beta) & \alpha \neq \beta, \\
    t e^{\alpha t} & \alpha = \beta,
    \end{cases}
\]
which is the standard bridge-expectation~\citep{HolmesRubin2002}
expression. The full integral over $[0, t_b]$ is a finite sum of
$\mathrm{HR}$ evaluations and closed-form coefficients in
$U, V, \lambda$.
\end{remark}

\subsubsection*{Optimization: Felsenstein-like pruning over node distributions}

The node distributions $\nu_v^s$ optimise the ELBO under a
Felsenstein-style upwards--downwards pass: each $\nu_v^s$ is the
geometric mean of (i) the upwards-propagated likelihood from
$\nu_v^s$'s subtree, (ii) the downwards-propagated prior from the
rest of the tree, and (iii) the per-branch forward and backward
propagators of $\hat Q^s$ on each adjacent branch. The pass is
$O(\text{edges} \cdot A^2)$ per outer ELBO step, dominated by the
per-branch matrix exponential. The rate $\hat Q^{s, *}_b$ is then
re-computed under the fresh node distributions and the loop iterates
to convergence; the ELBO is monotone increasing in each step.

\subsubsection*{Jensen gap and the Cohn upgrade as endpoints of an
adaptive refinement}

The constant-rate scheme of the previous subsection and the
inhomogeneous mean-field scheme of Cohn et al.~\citep{cohn2010} are
not separate algorithms but the $N = 1$ and $N \to \infty$ ends of a
single adaptive refinement. Subdivide a branch of length $t_b$ into
$N$ sub-intervals at times $0 = u_0 < u_1 < \cdots < u_N = t_b$ with
constant rate $\hat Q^s_{b, k}$ on each $[u_{k-1}, u_k]$, and a
variational node distribution $\nu^s_k$ at each intermediate
midpoint. The resulting variational law is a piecewise-constant CTMC
bridge: a piecewise-constant approximation to whatever time-varying
rate Cohn's scheme would use at the same fidelity. As $N \to \infty$
with mesh refining, piecewise-constant rates become dense in
continuous time-varying rates, so the variational family becomes
Cohn's and the ELBO converges monotonically up. At the optimum, the
$\nu^s_k$ are pinned by stationarity to the bridge marginals of the
surrounding piecewise-constant CTMC --- the explicit-midpoint
parameterisation is slack at the optimum.

The Jensen-gap diagnostic per segment is the integrated AM--GM
discrepancy of $Q^s$ against neighbour bridge marginals at the
segment's resolution; it is closed form via the same eigendecomposition
machinery and tells the algorithm where to refine. For typical
phylogenetic branches ($t_b \in [0.1, 1]$) most segments need
$N = 1$; high-coupling segments may benefit from $N = 2$ or $3$;
only pathological branches need $N \to \infty$.
Forced segmentation by composition-change events from the augmented
indel history (\S\ref{sec:gravestone-scfg}) composes cleanly with
this Jensen-driven refinement: subdivide first at every
composition-change event, then optionally $N$-refine within each
composition-constant segment.

\begin{remark}[Cohn's inhomogeneous fixed point as the $N \to \infty$ limit]
For a composition-constant segment requiring $N \to \infty$ refinement,
the limit is Cohn et al.'s full inhomogeneous fixed point: a
time-varying rate $\tilde Q^s(\cdot; u)$ obtained by alternating
forward sweeps $\dot q_s = q_s \tilde Q^s$ from $\delta_{x_p^s}$,
backward sweeps $\dot \rho_s = -\tilde Q^s \rho_s$ from
$\delta_{x_c^s}$, and rate updates
$\log \tilde Q^{s,*}(\cdot; u) = \mathbb{E}[\log Q^s(\cdot)]$
pointwise in $u$, with bridge marginals
$q^{\rm bridge}_s(x; u) \propto q_s(x; u)\, \rho_s(x; u)$.
The midpoint construction above is the discretisation of this
fixed point at mesh size $u_k - u_{k-1}$; the Cohn limit is recovered
as the mesh shrinks to zero. Cohn et al.~\citep[\S 7]{cohn2010}
also give the internal-node rule used by Felsenstein pruning under
the time-varying rate.
\end{remark}

This is the lower-bound construction referenced in the Discussion for
Potts components of size $> 2$: independent-coupling pairwise factorisation
gives the $N{=}1$ lower bound, midpoint augmentation extends it to a
formal ELBO, and $N \to \infty$ recovers the Cohn fixed point.
Classes whose Jensen gap remains large after refinement are
candidates for the $k$-clique cluster-variational upgrade of
Linzner \& Koeppl~\citep{linznerkoeppl2018}, or for fallback to
exact uniformization-based bridge sampling via Rao \&
Teh~\citep{raoteh2013}.

\subsection{Augmented indel histories via a time-indexed pair SCFG}
\label{sec:gravestone-scfg}

The class-level ELBO of \S\ref{sec:class-elbo} was derived under
fixed class composition over a constant-composition segment of a
branch. To compose with the tree we need the timed indel history on
each branch, including \emph{transient} insertions: under coupling,
an insertion that is born and then deleted on the same branch is not
visible in the alignment but does change the rate of every other
site in its DP class during its lifetime. The naive scheme of
sampling alignment-visible indel timings from their TKF92 prior is
inadequate for two distinct reasons: first, transient lineages are
missed; second, even for visible events, the conditional density
given the alignment depends on substitution evidence and is not what
the alignment marginal alone provides.

We resolve both issues by augmenting the latent state with
\emph{gravestones}. The augmentation does not change the model. It
changes the latent representation: every fragment that ever existed
during the branch --- alive at start, alive at end, deleted on the
branch, or transiently inserted then deleted --- is represented as
a labeled position in the augmented branch history. With gravestones
the alive count $N_a(s)$ is bounded by the visible count $N_v(s)$,
so the indel CTMC bridge has a finite state space and admits a
closed-form recursive sampler. The gravestone counts at internal
nodes of the tree become latent variables resampled jointly with the
rest of the augmented history.

\subsubsection*{The branching process and its grammar}

We first present the TKF91 case (single residue per fragment) where
the branching structure is cleanest, then read off the TKF92
expansion. Let $L(t)$ denote a single link observed $t$ time units
after its birth. Under TKF91 dynamics, $L(t)$ has lifetime
$\tau \sim \mathrm{Exp}(\mu)$, and during its lifetime spawns child
links at rate $\lambda$. Each child is itself an $L$ of the
appropriate residual age. The branching process is a
continuous-time Galton--Watson process expressible as a stochastic
context-free grammar with productions indexed by continuous time:
\begin{align*}
L(t) &\;\to\; A \cdot D(t,\, t)
& &\text{at weight } e^{-\mu t}, \\
L(t) &\;\to\; G_\tau \cdot D(t,\, \tau)
& &\text{at density } \mu e^{-\mu \tau}\, d\tau, \;\; \tau \in (0, t), \\
D(t,\, T) &\;\to\; L(t - u) \cdot D(t,\, T)
& &\text{at density } \lambda \, du, \;\; u \in (0, T), \\
D(t,\, T) &\;\to\; \varepsilon
& &\text{at weight } e^{-\lambda T}.
\end{align*}
The terminals $A$ and $G_\tau$ stand for an alive residue and a
gravestone residue that died at relative time $\tau$, respectively,
each carrying a DP key, a site class, and an amino-acid trajectory
over its lifetime as governed by \S\ref{sec:class-elbo}. The
non-terminal $D(t, T)$ generates a list of child links born during
a window of size $T$ inside an observation interval of size $t$.
The reading is: a link either survives to observation (probability
$e^{-\mu t}$) and can have spawned children at any time in
$(0, t)$, or it died at some $\tau \in (0, t)$ (density
$\mu e^{-\mu \tau}$) and can only have spawned children during
$(0, \tau)$. The empty-list weight $e^{-\lambda T}$ is the
probability of zero events from a Poisson process of rate $\lambda$
over a window of $T$.

\paragraph{TKF92 expansion.} The generalization to TKF92 replaces
each single-residue terminal with a fragment of geometrically-distributed
length, leaving the fragment-level branching structure entirely
unchanged:
\begin{align*}
A &\;\to\; \mathcal{R}^A \cdot A^*,
& A^* &\;\to\; \mathcal{R}^A \cdot A^* \;\;\text{at weight } r,
& A^* &\;\to\; \varepsilon \;\;\text{at weight } 1 - r, \\
G_\tau &\;\to\; \mathcal{R}^G_\tau \cdot G_\tau^*,
& G_\tau^* &\;\to\; \mathcal{R}^G_\tau \cdot G_\tau^* \;\;\text{at weight } r,
& G_\tau^* &\;\to\; \varepsilon \;\;\text{at weight } 1 - r,
\end{align*}
where $\mathcal{R}^A$ is a single alive residue and
$\mathcal{R}^G_\tau$ a single gravestone residue with shared death
time $\tau$. Fragment death is collective: all residues of a
fragment share its lifetime $[0, \tau]$ and die together at $\tau$.
Key-DP labels $z_s$ and class-DP labels $c_s$ are drawn
independently per residue at fragment birth from their respective
Chinese restaurant predictives, even within a single fragment, so a
fragment can carry residues of several key classes and several site
classes. The expansion is invisible to the indel skeleton --- the
bivariate Riccati of the next subsection lives at the fragment level
--- but enters the substitution likelihood and the DP-class
composition through the per-residue payload.

Pairing this with an ancestor sequence yields the pair version: each
ancestor fragment present at branch start is an independent root of
the recursion above, with $t = t_b$ the branch length; insertions
appearing in the children list have no ancestor counterpart and
carry only a descendant payload. The pair grammar generates an
aligned pair of fragment sequences (with alive/gravestone labels on
the descendant side) jointly with the parentage tree.

\subsubsection*{Inside generating function and the bivariate Riccati}

Let $f_t(z, w) = \mathbb{E}[z^{N_a(t)} w^{N_g(t)}]$ where $N_a, N_g$
count alive and gravestone descendants of one root fragment of age
$t$. The inside generating function of the grammar satisfies the
Volterra fixed point
\[
f_t(z, w)
\;=\; z\, e^{-\mu t}\, \exp\!\Bigl(\lambda \!\int_0^t (f_s(z, w) - 1)\, ds\Bigr)
\;+\; w\!\int_0^t \mu e^{-\mu \tau}\, \exp\!\Bigl(\lambda \!\int_{t-\tau}^t (f_s(z, w) - 1)\, ds\Bigr)\, d\tau,
\]
which is equivalent to the forward equation
\[
\partial_t f_t(z, w) \;=\;
\bigl[\lambda z^2 - (\lambda + \mu) z + \mu w\bigr] \, \partial_z f_t(z, w),
\qquad f_0(z, w) = z.
\]
The right-hand side is quadratic in $z$, so the characteristic
equation $\dot z = -[\lambda z^2 - (\lambda + \mu) z + \mu w]$ is
Riccati. Define
\[
\Delta(w) \;=\; \sqrt{(\lambda - \mu)^2 + 4 \lambda \mu (1 - w)},
\qquad
z_\pm(w) \;=\; \frac{(\lambda + \mu) \pm \Delta(w)}{2 \lambda}.
\]
The Riccati flow
$(z(t) - z_+) / (z(t) - z_-) = (z(0) - z_+)/(z(0) - z_-) \cdot e^{-\Delta(w) t}$
inverts to give $f_t(z^*, w)$ in closed form, hence closed-form
pointwise transition probabilities
$P\!\bigl((N_a, N_g)(t)\,\big|\,(N_a, N_g)(0)\bigr)$ by coefficient
extraction. Setting $w = 1$ recovers the standard linear
birth-death PGF for $N_a$ alone.

\subsubsection*{Recursive traceback sampler}

Given an augmented history with endpoint counts
$(a_0, g_0) = (N_a(0), N_g(0))$ and
$(a^*, g^*) = (N_a(t_b), N_g(t_b))$, the within-branch indel
history is sampled by recursive midpoint traceback on the bivariate
process $(N_a, N_g)$. At each level of the recursion, sample a
midpoint state from
\[
P\!\bigl((a_m, g_m) \,\big|\, (a_0, g_0), (a^*, g^*)\bigr)
\;\propto\;
P\!\bigl((a_0, g_0) \to (a_m, g_m);\, t_b/2\bigr) \cdot
P\!\bigl((a_m, g_m) \to (a^*, g^*);\, t_b/2\bigr),
\]
both factors closed form via the Riccati. The midpoint state space
is finite since $a_m, g_m \leq g^* + a^*$. Recurse on each half.
The base case is reached when the residual subproblem has at most a
small number of events (births and deaths), at which point the
closed-form list of valid event interleavings is enumerable
directly and event times are sampled from the corresponding
hypoexponential simplex density. Total cost is
$O((b + d) \log(b + d))$ per branch where $b, d$ are the branch
totals of births and deaths.

After event times are placed, parentage is sampled forward in time:
at each birth, a uniform alive fragment is chosen as the parent; at
each death, a uniform alive fragment is killed. Parentage uniformity
is the standard exchangeability of the linear Galton--Watson tree.

\subsubsection*{Inserted fragment and residue distributions}

When a child fragment is added at time $u$ to a DP class $C$ that
already contains alive fragments, the inserted residues are drawn
from the conditional Boltzmann
\[
P\!\bigl(x_s(u^+) \,\big|\, x_{C \setminus s}(u^-)\bigr)
\;\propto\;
\pi^{(c_s)}\!\bigl(x_s(u^+)\bigr)\, \exp\!\Big(-\!\!\sum_{s' \in C \setminus s}\!\! H_{c_s c_{s'}}\!\bigl(x_s(u^+),\, x_{s'}(u^-)\bigr)\Big),
\]
which is exactly normalizable on the alphabet $\mathcal{A}$ even
when the joint stationary on $C$ is intractable. This is the proper
coupled generalization of the $\mathrm{Cat}(\pi)$ inserted-residue
rule of standard TKF92.

\subsubsection*{Composition with substitution and with the rest of
the tree}

The traceback gives a piecewise-constant DP-class composition on
each branch, with segment boundaries at every (alive or gravestone)
indel event. The variational ELBO of \S\ref{sec:class-elbo} applies
to each segment as written, with the segment's class composition
and duration. Bridge expectations accumulate over
segments within each branch and across branches.

Internal-node augmented states $(N_a, N_g)$-valued are sampled
jointly across the tree. The standard upwards-downwards
(Felsenstein-style) recursion applies because the closed-form
per-branch transition kernel from the Riccati composes across
branches: the marginal at each internal node, conditional on the
rest of the tree, is a closed-form pointwise distribution
computable in $O(1)$ given the per-branch kernel evaluations. The
augmentation thus extends to a full tree-level closed-form sampler.

\subsubsection*{Marginalizing the augmentation}

Gravestones are an augmentation of the latent state, not of the
data. The marginal likelihood of the observed alignment under the
model is recovered by summing over gravestone configurations at
every internal node and along every branch. The recursive traceback
samples the augmented state, and the variational ELBO of
\S\ref{sec:class-elbo} applied segment-wise on the augmented
branches gives the conditional substitution likelihood. No other
approximation enters at the indel level: the SCFG is the exact
representation of TKF92's fragment-level dynamics, and the
gravestone augmentation is what makes its inside-outside computable.

\begin{remark}[Coalescent priors; ARG augmentation as a parallel construction]
The per-branch SCFG depends on branch length only through $t_b$, so
any prior on the tree --- including a coalescent prior on branch
times --- composes with the gravestone-augmented sampler without
modification. This is a free generalization that supports joint
indel-aware coalescent inference on protein-coding regions, which
existing coalescent frameworks typically handle by treating indels
as missing data. Separately, the structural mechanism --- augmenting
an unbounded latent process with non-observable entities so that an
observed count provides a state-space bound --- carries over to the
ancestral recombination graph: non-ancestral lineages in the ARG
play the role of gravestones, bounding the per-genome-position bridge
state space and admitting (we conjecture) a parallel closed-form
recursive sampler.
\end{remark}

\subsection{Posterior sampling and parameter learning}
\label{sec:tkfdp-svi}

We turn to the inverse problem: given a corpus of MSAs
$\{X_n\}_{n=1}^N$ with associated trees $\{T_n\}$ and fixed
alignment paths, learn the corpus-level substitution hyperparameters
\[
\theta = \bigl(S,\, \alpha_z,\, \alpha_c,\, \alpha_H,\, a_\eta, b_\eta,\,
\kappa_\pi, \bar\pi,\, \{\mu_{kl}, \tau_{kl}\}_{k \leq l}\bigr),
\]
namely the shared exchangeability $S$, the three DP concentrations
$(\alpha_z, \alpha_c, \alpha_H)$, the per-site rate prior
$(a_\eta, b_\eta)$, and the base-measure parameters for $G_0$ on
$\pi^{(c)}$ and for $G_0^H$ on the entries of each Potts atom
$H_h$. The per-site rate multipliers $\{\eta_s\}$, the per-class
profiles $\{\pi^{(c)}\}$, the materialized Potts atoms $\{H_h\}$,
and the class-pair Potts assignments $\{h_{cc'}\}$ are no longer
parameters but \emph{latent} draws from the base measures; their
posteriors are inferred per site, per class, per atom, and per
class-pair respectively, with the corpus-level parameters governing
the priors. Indel parameters $(\lambda, \mu, r)$ decouple cleanly at
the parameter level via the alignment--substitution factorization
of \S\ref{sec:tkfdp-model} and are learned from the alignment
marginal alone by standard TKF92 EM; we focus on the substitution
side. The latent variables on each MSA are the key-DP labels
$\{z_s\}$, the site classes $\{c_s\}$, the per-site rates
$\{\eta_s\}$, the within-class amino-acid trajectories
$\{\sigma_C\}$, the unobserved internal-node states, the Potts-atom
assignments $\{h_{cc'}\}$, and (per \S\ref{sec:gravestone-scfg}) the
gravestone-augmented indel histories on every branch --- comprising
the alive and gravestone counts $(N_a, N_g)$ at every internal
node, plus the parentage tree and event times within each branch.

\subsubsection*{Variational EM with bridge-expectation sufficient
statistics}

For singleton key-DP classes, $Q^s$ reduces to F81 with
class-specific equilibrium, shared exchangeability, and per-site
rate. The bridge expectations~\citep{HolmesRubin2002} --- expected transition counts $N^{(b,s)}_{xy}$ and
dwell times $T^{(b,s)}_x$ along each branch $b$, conditioned on the
endpoint amino acids --- are computable in closed form by
integrating $\int_0^{t_b} e^{Q u} A\, e^{Q(t_b - u)}\,du$ in the
eigenspace of the rate matrix. Aggregated appropriately they give
closed-form posterior updates on every continuous latent on the
substitution side:
\begin{itemize}
    \item $\pi^{(c)}$ has a Dirichlet posterior under the
      $\mathrm{Dirichlet}(\kappa_\pi \bar\pi)$ prior, via the
      secret-destination augmentation of
      \S\ref{sec:base-measure-conj} below.
    \item $\eta_s$ has a Gamma posterior under the
      $\mathrm{Gamma}(a_\eta, b_\eta)$ prior; both shape and rate
      updates are immediate from the bridge expectations.
      The marginal site likelihood
      $\int \mathrm{Gamma}(a_\eta, b_\eta) \cdot \mathrm{Poisson}(N_s^{\mathrm{acc}} \mid \eta_s \tilde T_s)\, d\eta_s$
      is closed-form Negative-Binomial.
    \item Each materialized Potts atom $H_h$ has a Gaussian
      posterior under the $\mathcal{N}(\mu_{kl}, \tau_{kl}^{-1})$
      prior, via the Gaussian (Hessian-based) Laplace approximation
      of the path-DCA likelihood at the current iterate.
\end{itemize}
The shared exchangeability $S$ aggregates over all classes as in
standard GTR EM. Site-class assignments $c_s$ and Potts-atom
assignments $h_{cc'}$ are latent in the CRP sense, integrated over
via the Gibbs sweeps below.

For multi-site DP classes the joint generator on
$\mathcal{A}^{|C|}$ has no usable eigendecomposition, and
the bridge expectations cannot be applied to the coupled bridge directly. The
variational construction of \S\ref{sec:class-elbo} is the
substitute: at the variational fixed point each $s \in C$ has an
inhomogeneous single-site bridge with rate $\tilde Q^{s,*}$ on each
constant-composition segment of the branch
(\S\ref{sec:gravestone-scfg}), and the bridge expectations apply to this
bridge in its eigenspace. The integral identity goes through with
$Q$ replaced by $\tilde Q^{s,*}(u)$ (piecewise-constant on a time
grid within each segment), yielding per-site sufficient statistics
$\tilde N^{(b,s)}_{xy}$, $\tilde T^{(b,s)}_x$ as well as per-pair
co-occupancy times
\[
\tilde T^{(b, s, s')}_{xy} \;=\; \int_0^{t_b} q_s(x; u)\, q_{s'}(y; u)\, du,
\]
the latter immediate from the variational independence of single-site
bridges given endpoints.

The envelope theorem ensures that, evaluated at the variational
fixed point, the gradient of the ELBO with respect to $\theta$ has
no contribution from the implicit $\theta$-dependence of
$\tilde Q^{s,*}$:
\[
\nabla_\theta \mathcal{L}
\;=\; \mathbb{E}_\mathbb{Q}\!\bigl[\nabla_\theta \log \tfrac{d\mathbb{P}}{d\mathbb{Q}}\bigr]
\;=\; \mathbb{E}_\mathbb{Q}\!\bigl[\nabla_\theta \log d\mathbb{P}\bigr].
\]
For the GTR-Potts generator this gradient has three structurally
distinct pieces. The exchangeability gradient $\nabla_S \mathcal{L}$
is a linear combination of the single-site $\tilde N^{(b,s)}_{xy}$
counts as in standard GTR EM. The frequency gradient
$\nabla_{\pi^{(c)}} \mathcal{L}$ aggregates over sites with
$c_s = c$ and is again single-site. The coupling-tensor gradient is
a path-space DCA M-step,
\[
\nabla_{H_{cc'}(x, x')} \mathcal{L}
\;\propto\; \!\!\sum_{n, b}\!\!\sum_{\substack{s, s' \in C(n) \\ c_s = c,\, c_{s'} = c'}}\!\!\!\Bigl[\,\bigl(\text{expected } x_s = x,\, x_{s'} = x' \text{ co-occupancy under } \mathbb{P}\bigr) - \bigl(\text{same under } \mathbb{Q}\bigr)\,\Bigr],
\]
expressible in the same bridge-expectation format using the products
$q_s(x; u) q_{s'}(x'; u)$ for the variational dwell-time terms and
contributions from each variational jump intensity for the
transition terms.

\subsubsection*{MCMC over the discrete latents}

The combinatorial latents $\{z_s\}$, $\{c_s\}$, and $\{h_{cc'}\}$
admit straightforward Gibbs sampling. All three partitions are
integrated over their respective stick-breaking weights via Chinese
restaurant predictives. For the key DP,
\[
P(z_s = k \mid z_{-s}, \alpha_z)
\;\propto\;
\begin{cases}
|C_k^{(-s)}| \cdot \mathcal{L}_{\text{class}}(C_k \cup \{s\}) / \mathcal{L}_{\text{class}}(C_k)
& \text{(existing key class)}, \\
\alpha_z \cdot \mathcal{L}_{\text{class}}(\{s\})
& \text{(new singleton)},
\end{cases}
\]
where $\mathcal{L}_{\text{class}}$ is the per-branch-aggregated
class-level variational ELBO from \S\ref{sec:class-elbo} plus the
contribution of the rest of the tree via Felsenstein pruning.
Site-class updates use the class DP predictive,
\[
P(c_s = c \mid c_{-s}, \alpha_c, \theta)
\;\propto\;
\begin{cases}
n_c^{(-s)} \cdot \mathcal{L}_{\text{site}}(s; \pi^{(c)})
& \text{(existing class, profile $\pi^{(c)}$)}, \\
\alpha_c \cdot \int \mathcal{L}_{\text{site}}(s; \pi')\, G_0(d\pi')
& \text{(new class)},
\end{cases}
\]
with the new-class integral closed form by the
Dirichlet--multinomial conjugacy of
\S\ref{sec:base-measure-conj}. The Potts-atom assignment
$h_{cc'}$ for an unordered class-pair $\{c, c'\}$ is itself drawn
by CRP-Gibbs,
\[
P(h_{cc'} = h \mid \text{rest}, \alpha_H)
\;\propto\;
\begin{cases}
m_h^{(-cc')} \cdot \mathcal{L}_{\text{pair}}(\{c, c'\};\, H_h)
& \text{(existing atom)}, \\
\alpha_H \cdot \int \mathcal{L}_{\text{pair}}(\{c, c'\};\, H')\, G_0^H(dH')
& \text{(new atom)},
\end{cases}
\]
where $\mathcal{L}_{\text{pair}}$ is the path-DCA contribution from
co-occurrences of classes $c$ and $c'$ across the corpus, and the
new-atom integral is approximated by the Laplace scheme of
\S\ref{sec:base-measure-conj}.

Split--merge proposals of Jain--Neal type~\citep{jainneal2004} are
essential for mixing on all three partitions, with acceptance ratios
in terms of class-level (for $z$), site-aggregated (for $c$), or
class-pair-aggregated (for $h$) ELBO ratios. All three concentrations
$\alpha_z, \alpha_c, \alpha_H$ admit closed-form Gamma-prior
posteriors given the respective partitions via the auxiliary-variable
scheme of Escobar \& West~\citep{escobarwest1995}.

\subsubsection*{Truncated stick-breaking on Potts atoms}

The Potts-atom CRP admits a truncated stick-breaking (TSB)
alternative that is cheaper per sweep, requires no new-atom Laplace
branch inside the Gibbs step, and removes the empty-atom bookkeeping
of the CRP. The construction mirrors Ishwaran \&
James~\citep{ishwaranjames2001} and Blei \&
Jordan~\citep{bleijordan2006} as applied to the site-class DP,
transposed to the Potts-atom DP.

Fix a truncation level $K_H^{\max}$. The natural choice is
$K_H^{\max} = K_c(K_c + 1)/2$, the total number of unordered
class-pairs --- with this bound the truncation is exactly tight
(every class-pair could in principle have its own atom). Replace
the CRP-Gibbs sweep over $\{h_{cc'}\}$ with the following two-step
alternation:

\begin{enumerate}
\item \textbf{Beta CAVI / Gibbs on stick proportions.} Maintain
  stick variables $\beta_h \sim \mathrm{Beta}(1, \alpha_H)$ for
  $h < K_H^{\max}$, with $\beta_{K_H^{\max}} = 1$ and
  $\rho_h = \beta_h \prod_{h' < h} (1 - \beta_{h'})$. Given the
  current per-atom counts $m_h = \#\{(c, c') : h_{cc'} = h\}$, the
  conditional posterior on $\beta_h$ is the closed-form Beta
\[
    \beta_h \mid m, \alpha_H \;\sim\;
    \mathrm{Beta}\!\Bigl(1 + m_h,\; \alpha_H + \!\!\sum_{h' > h}\!\! m_{h'}\Bigr).
\]
Use either the posterior mean (CAVI) or a Beta sample (Gibbs).

\item \textbf{Categorical resample on $\{h_{cc'}\}$.} For each
  unordered class-pair $\{c, c'\}$, draw the atom assignment from
  the closed-form Categorical
\[
    P(h_{cc'} = h \mid \text{rest}) \;\propto\;
    \rho_h \cdot \mathcal{L}_{\text{pair}}(\{c, c'\};\, H_h),
\]
where $\mathcal{L}_{\text{pair}}$ is the path-DCA contribution from
co-occurrences of classes $c$ and $c'$ across the corpus, evaluated
at the current materialized atom $H_h$.
\end{enumerate}

\begin{remark}[Bijective initialization]
Initialization at $K_H^{\max} = K_c(K_c + 1)/2$ should set each
unordered class-pair to a \emph{distinct} atom drawn from $G_0^H$,
not collapse all class-pairs onto a single shared atom. The latter
is a chicken-and-egg failure mode: only the populated atom receives
data through the Laplace M-step, so the others stay at the prior;
the next TSB resample therefore prefers the populated atom for all
pairs (its likelihood dominates the others' prior-only score), and
the system collapses. The bijective init guarantees every atom
carries at least one class-pair's data through the first Laplace
M-step.
\end{remark}

\subsubsection*{Locality of DP updates under the SCFG augmentation}

When the augmented indel history is resampled
(\S\ref{sec:gravestone-scfg}), new gravestone fragments appear with
new residues that need DP keys, and old gravestones disappear.
Naively this would suggest a full DP-key resample after every
augmentation step, which would be prohibitive. Exchangeability of
the CRP and the segment-wise structure of the variational ELBO
together remove almost all of this cost.

The decomposition is two-step within a single sweep. The
\emph{indel-skeleton step} resamples the gravestone-augmented history
on every branch by the SCFG traceback of \S\ref{sec:gravestone-scfg}
without touching DP keys; cost $O((b + d) \log(b + d))$ per branch,
independent of DP state. The \emph{gravestone-key step} then
assigns DP keys to each new residue $s$ on a gravestone fragment
with lifetime $[u, v]$ on branch $b$, using the local CRP
predictive over existing classes plus a singleton, evaluated against
$\mathcal{L}_{\text{seg}}$ on segments of $b$ within the gravestone's
lifetime. Adding $s$ to class $C_k$ changes that class's variational
fixed point only on those segments, costing $O(|C_k| \cdot A^3)$
per fixed-point iteration per touched segment. Per gravestone, the
work is one Categorical draw with support equal to the existing
classes plus a singleton, plus one local fixed-point re-solve on
the affected segments.

\subsubsection*{Closed-form base-measure integrals via complete-data augmentation}
\label{sec:base-measure-conj}

The base-measure integrals
\[
    \int \mathcal{L}_{\text{site}}(s; \pi')\, G_0(d\pi'), \qquad
    \int \mathcal{L}_{\text{pair}}(\{c, c'\};\, H')\, G_0^H(dH'),
\]
that appear in the new-class and new-atom branches of the CRP
predictives, and the analogous corpus-level marginals over
$\eta_s$, all admit closed-form approximations under complete-data
augmentation.

\paragraph{Secret-destination augmentation for $\pi^{(c)}$.} The
F81-form rate factors as
$Q^s_{xy} = \eta_s \cdot S_{xy} \cdot \pi^{(c_s)}(y) \cdot \exp(-\tfrac{1}{2}\Delta H_s)$,
which decomposes the dynamics into an $S$-driven Poisson proposal
clock at rate $\eta_s S_{xy}$ from state $x$ to destination $y$,
modulated by a Bernoulli filter that accepts the proposal with
probability $\pi^{(c_s)}(y) \exp(-\tfrac{1}{2}\Delta H_s)$. To
recover full multinomial structure on $\pi^{(c)}$ we augment each
silent failure of a proposal of $y$ with a categorical \emph{secret
destination} $j \neq y$, drawn from $\pi^{(c)}(j)/(1 - \pi^{(c)}(y))$
on the simplex. Each proposal then casts exactly one categorical
vote: the destination is $y$ with probability $\pi^{(c)}(y)$ on
acceptance, and a secret $j$ with probability $\pi^{(c)}(j)$ on
rejection. Marginally over the proposal, the vote is
$\mathrm{Categorical}(\pi^{(c)})$, so the augmented count vector
$N^{(c)} \in \mathbb{Z}_{\geq 0}^A$ satisfies
\[
    N^{(c)} \mid \pi^{(c)},\, \text{augmentation} \;\sim\;
    \mathrm{Multinomial}\bigl(N^{(c)}_{\text{total}},\, \pi^{(c)}\bigr),
\]
and the posterior is the strict closed-form Dirichlet
\[
    \pi^{(c)} \mid N^{(c)} \;\sim\;
    \mathrm{Dirichlet}\bigl(\kappa_\pi \bar\pi + N^{(c)}\bigr).
\]
In the EM/VBEM setting the augmentation is replaced by its
conditional expectation: each proposal contributes $\pi^{(c)}(y)$ to
$E[N^{(c)}_y]$ regardless of which destination was proposed. The
Dirichlet update is then a fixed-point iteration whose limit is the
conjugate posterior under the conditional augmentation. The
new-class integral $\int \mathcal{L}_{\text{site}}(s; \pi')\, G_0(d\pi')$
is the analytic ratio of multivariate Beta functions
$B(\kappa_\pi \bar\pi + N^{(c)}) / B(\kappa_\pi \bar\pi)$.

\paragraph{Gamma--Poisson conjugacy for $\eta_s$.} The F81
substitution-count likelihood at site $s$ is itself Gamma-conjugate
in $\eta_s$ without any need for the secret-destination
augmentation. The expected substitution count along the tree is
Poisson with rate $\eta_s \cdot \tilde T_s$, where
\[
    \tilde T_s \;=\; \sum_b \sum_x T^{(b,s)}_x \cdot
    \sum_{y \neq x} S_{xy} \, \pi^{(c_s)}(y)
\]
is the $\pi$-weighted dwell-time integral. Under the
$\mathrm{Gamma}(a_\eta, b_\eta)$ prior the conjugate posterior is
closed form
\[
    \eta_s \mid \mathrm{path},\, \pi^{(c_s)} \;\sim\;
    \mathrm{Gamma}\!\bigl(a_\eta + N_s^{\mathrm{acc}},\; b_\eta + \tilde T_s\bigr),
\]
with $N_s^{\mathrm{acc}} = \sum_b \sum_{x \neq y} N^{(b,s)}_{xy}$
the total observed substitution count, and the per-site marginal
likelihood is closed-form Negative-Binomial.

\paragraph{Laplace for $H_h$.} Each Potts atom is real-valued and
the path-DCA likelihood is locally Gaussian in $H_h$ to second
order. We approximate
\[
    \log p_{\mathrm{path-DCA}}(\mathrm{data} \mid H_h)
    \;\approx\; \log p_{\mathrm{path-DCA}}(\mathrm{data} \mid \hat H_h)
    \;-\; \tfrac{1}{2} (H_h - \hat H_h)^\top \Lambda_h (H_h - \hat H_h),
\]
where $\hat H_h$ is the MAP under the current iterate, found by a
few Newton or natural-gradient steps starting from a chosen seed,
and $\Lambda_h = -\nabla^2_{H_h} \log p_{\mathrm{path-DCA}}$ is the
Hessian at $\hat H_h$. Combined with the
$\mathcal{N}(\mu_{kl}, \tau_{kl}^{-1})$ prior on each entry, the
posterior is the closed-form Gaussian
\[
    H_h \mid \mathrm{data} \;\sim\;
    \mathcal{N}\!\bigl(\hat H_h^{\mathrm{post}},\, (\mathrm{diag}(\tau_{kl}) + \Lambda_h)^{-1}\bigr).
\]
The Laplace estimate of the new-atom integral is the standard
$\log p_{\mathrm{path-DCA}}(\hat H_h) + \log G_0^H(\hat H_h) + \tfrac{d}{2}\log(2\pi) - \tfrac{1}{2}\log\det(\mathrm{diag}(\tau_{kl}) + \Lambda_h)$
with $d = A(A+1)/2$ the symmetric-slice dimension.

\paragraph{Multi-seed Laplace mixture for multimodal posteriors.}
The Laplace approximation collapses around a single mode and
underestimates posterior mass elsewhere; for atoms that admit
multiple plausible coupling patterns (e.g., charge-complementary
vs.\ hydrophobic-hydrophobic minima) this is a genuine approximation
error. We mitigate by running the few-step optimizer from a small
number $K_{\mathrm{seed}}$ of distinct seeds, obtaining
$K_{\mathrm{seed}}$ Laplace components, and combining them as a
weighted Gaussian mixture.

\subsubsection*{The recommended hybrid}

The natural hybrid is:
\begin{enumerate}
    \item \textbf{Inner variational E-step.} For each MSA, branch
      and DP class, segment the branch by the current sampled
      augmented indel history and solve the geometric-mean fixed
      point of \S\ref{sec:class-elbo} on each segment to obtain
      $\{\tilde Q^{s,*}\}$ and the per-class ELBO. No parameter or
      structure updates here.
    \item \textbf{MCMC over discrete and augmentation latents.}
      Resample augmented indel histories on every branch by recursive
      traceback on the time-indexed pair SCFG of
      \S\ref{sec:gravestone-scfg}, with closed-form midpoint
      marginals from the bivariate Riccati and Felsenstein-style
      upwards--downwards composition for internal-node augmentation
      counts. Gibbs sweep over $\{z_s\}$, $\{c_s\}$, and
      $\{h_{cc'}\}$ using the class-level ELBOs from step~1 as
      evidence; periodic Jain--Neal split--merge proposals on each
      partition.
    \item \textbf{Stochastic M-step.} Compute bridge-expectation
      sufficient statistics $\tilde N$, $\tilde T$,
      $\tilde T_{\mathrm{pair}}$ from the variational single-site
      bridges, summed over augmented constant-composition segments
      within each branch, over a minibatch of MSAs. Take a stochastic
      gradient step on the corpus-level hyperparameters
      $\theta = (S, \alpha_z, \alpha_c, \alpha_H, a_\eta, b_\eta, \kappa_\pi, \bar\pi, \{\mu_{kl}, \tau_{kl}\})$.
      Per-class $\pi^{(c)}$ updates by its closed-form Dirichlet
      posterior under the secret-destination augmentation; per-site
      $\eta_s$ marginalizes in closed form via Gamma--Poisson
      conjugacy; per-atom $H_h$ updates by Laplace with multi-seed
      mixture against the prior $G_0^H$. The three DP concentrations
      update by the Escobar--West auxiliary-variable scheme.
    \item \textbf{Iterate.} The inner variational fixed point in
      step~1 is re-solved cheaply after each step because the
      previous solution is a warm start.
\end{enumerate}

This factors the overall problem into the three regimes where each
tool is strongest: variational for the continuous-time substitution
dynamics where it admits closed-form rates and bridge-expectation
sufficient statistics; MCMC for the discrete combinatorial latents
where it mixes well and avoids mean-field pathologies; stochastic
gradient for the continuous parameters $\theta$ where it scales to
corpus-level training. In SVI form the M-step is over a minibatch
of MSAs at each iteration and amortizes naturally over the corpus.

\subsubsection*{Counts-tensor representation of the training corpus}

Per family we precompute a counts tensor
$(\text{cherry} \times \text{column} \times \text{AA-pair} \times \text{branch length})$,
packed as int8, that decouples inference from raw MSA parsing and
lets one outer SVI iteration scale to $10^3$--$10^4$ \Pfam\ families.
Branch lengths are quantized to $n_t \approx 50$ geometrically spaced
bins between $\tau_{\min}$ and $\tau_{\max}$; geometric spacing
exploits that $\exp(Q\tau)$ varies fastest at small $\tau$, giving a
$\sim 5\times$ cache reduction over linear quantization at no
measurable accuracy cost. Per-class single-site log-transition
matrices $(K_c, n_t, A^2)$ are precomputed once per outer iteration
and shared across all families.

\subsection{Pairwise alignment postprocessing}
\label{sec:postprocessing}

Multiple-sequence alignment tools of the FSA
family~\citep{bradley2009fsa} (and other consistency-based schemes)
operate on pairwise residue posterior matrices
$Q^{(X, Y)}_{ij} = P(X_i \sim Y_j \mid X, Y)$ produced by per-pair
forward--backward passes through a Pair HMM. The TKF-DP postprocessing
target is the joint-marginal alignment-and-partition likelihood
described next, in which an alignment $A$ together with a
size-$\{1, 2\}$ partition $E$ of its Match cells contributes a
product of per-block substitution factors with no class or
rate-multiplier latent left explicit.

\begin{remark}[Historical note: deprecated first-order correction]
\label{rem:deprecated-mean-field}
An early formulation of TKF-DP postprocessing expressed the
coupling as a first-order mean-field correction
$\log Q'_{ij} - \log Q_{ij} \approx \varepsilon \sum_{(i', j')}
  Q_{i' j'} [M(i, j; i', j'; t) - 1]$
to the baseline single-site posterior, with $\varepsilon$ a
per-pair partner prior under the size-$\{1, 2\}$ Ewens
distribution and $M$ a column-class-marginalized Potts boost. That
correction is a first-order expansion of the joint marginal in
the partner-prior strength and is now considered a heuristic
stepping-stone; the modern infinite Pair HMM
(\S\ref{sec:infinite-hmm} of Appendix~\ref{app:phmm}) treats the
same coupling exactly via the joint-marginal target below, with
class assignments and rate multipliers integrated analytically
rather than re-marginalized at every column.
\end{remark}

\subsubsection*{The unified joint-marginal target}

Let $\theta_{\mathrm{indel}} = (\lambda, \mu, r)$ be the TKF92
indel parameters and $\alpha_z$ the Ewens partition concentration.
Write $\mathrm{Match}(A)$ for the Match cells of an alignment $A$,
and let $E$ be a set of unordered Match-cell edges (a partition of
$\mathrm{Match}(A)$ into singletons of size~1 and doublets of
size~2; we let $V(E) \subset \mathrm{Match}(A)$ denote the doublet
endpoints). The TKF-DP joint marginal at branch length $t$ is
\begin{equation}
\label{eq:tkfdp-joint-marginal}
P(X, Y; t) \;=\; \sum_A \pi_{\TKF92}(A \mid t, \theta_{\mathrm{indel}})
  \sum_E \pi_{\mathrm{Ewens}}(E \mid \alpha_z, |\mathrm{Match}(A)|)
  \!\!\!\prod_{(i, j) \in \mathrm{Match}(A) \setminus V(E)}\!\!\!\!\!\!
    P_{\mathrm{singlet}}(x_i, y_j; t)
  \prod_{e = ((i, j), (k, l)) \in E}\!\!\!\!\!
    P_{\mathrm{doublet}}\bigl((x_i, x_k), (y_j, y_l); t\bigr),
\end{equation}
with the per-block primitives
\begin{align}
P_{\mathrm{singlet}}(a, b; t)
  \;&=\; \sum_c \pi_c \,\, \pi^{(c)}(a) \,\, P_c(a \to b;\, t \cdot \eta),
  \label{eq:p-singlet} \\
P_{\mathrm{doublet}}\bigl((a, c), (b, d); t\bigr)
  \;&=\; \sum_{c_1, c_2} \pi_c(c_1) \, \pi_c(c_2)
  \,\, \pi_{\mathrm{joint}}^{c_1, c_2}(a, c)
  \,\, P_{\mathrm{joint}}^{c_1, c_2}\!\bigl((a, c) \to (b, d);\, t, \eta_1, \eta_2\bigr).
  \label{eq:p-doublet}
\end{align}
The doublet endpoint convention (matching the boost API of the
released code) is that $(a, b)$ are the amino acids at the left
endpoint $(i, j)$ and $(c, d)$ at the right endpoint $(k, l)$:
$a = x_i$, $b = y_j$, $c = x_k$, $d = y_l$. Indices $c_1, c_2$ run
over site classes and should not be confused with the
endpoint-amino-acid index $c$.

The empirical class prior $\pi_c$ is the per-class column count
from training (\emph{not} uniform $1 / K_c$); the per-class profile
$\pi^{(c)}(\cdot)$ is the trained \texttt{pi\_class}; the
class-conditional transition is
\begin{equation}
\label{eq:p-class-cond}
P_c(a \to b; t) \;=\; [\exp(Q_c \cdot t)]_{a, b},
\qquad
Q_c \;=\; (S_{\mathrm{LG08}} - \mathrm{diag}\, S_{\mathrm{LG08}})
          \cdot \pi^{(c)}\![\,\cdot\,]^\top,
\end{equation}
the standard F81-form GTR generator with exchangeability
$S_{\mathrm{LG08}}$~\citep{LG08} and class-specific stationary
$\pi^{(c)}$. The joint stationary
$\pi_{\mathrm{joint}}^{c_1, c_2}(a, c)$ at a coupled cell-pair is
the joint stationary of a Potts atom $H = \mathtt{atoms}[\mathtt{assignments}[c_1, c_2]]$
at the trained checkpoint (the trained
\texttt{potts\_dp.atoms} indexed by \texttt{potts\_dp.assignments}),
and $P_{\mathrm{joint}}^{c_1, c_2}$ is its time-evolved transition
$\exp(Q_{\mathrm{joint}}^{c_1, c_2}(H) \cdot t)$ built by
\texttt{build\_joint\_Q\_pair}$(H, \pi_a, \pi_b, S, \eta_{\mathrm{pair}} = (\eta_1, \eta_2))$
as described in \S\ref{sec:gtr-potts}.

\paragraph{Pair-background convention for the joint generator.}
Two natural choices for the pair background $(\pi_a, \pi_b)$
arise. The first is to use the LG08 single-site equilibrium
$\pi_a = \pi_b = \pi^{\mathrm{LG08}}$ for every class-pair
$(c_1, c_2)$. This is the \emph{canonical} convention for the
released $K{=}4$ EM-warmup checkpoint: the trained Potts atoms
were fit as deviations of the joint stationary from the
LG08-pair-background reference, so inference under that checkpoint
must use the same convention for the joint stationary and joint
generator to match training. The second choice,
$\pi_a = \pi^{(c_1)}$, $\pi_b = \pi^{(c_2)}$, uses the per-class
profiles as the pair background; this is the structurally
``cleaner'' choice for future model variants in which the Potts
atoms are retrained against the per-class background, but it is
\emph{not} consistent with the released checkpoint.

\paragraph{Discrete-gamma rate-multiplier integration.}
The per-site rate multiplier $\eta_s \sim \mathrm{Gamma}(a_\eta, b_\eta)$
of \S\ref{sec:gtr-potts} is integrated analytically by the discrete-gamma
quadrature of Yang~\citep{Yang1994a}: $K_r$ equiprobable bin
medians $\eta_1, \ldots, \eta_{K_r}$ are taken from the
$\mathrm{Gamma}(a_\eta, b_\eta)$ prior with uniform weight
$1 / K_r$, and the per-block primitive of
equation~\ref{eq:p-singlet} (resp.~\ref{eq:p-doublet}) is averaged
over rate-multiplier draws via
$P_{\mathrm{singlet}}(a, b; t) = K_r^{-1} \sum_{r = 1}^{K_r}
  P_{\mathrm{singlet}}(a, b; t \cdot \eta_r)$
(resp.\ a double sum over $(\eta_1, \eta_2)$ for the doublet).
The Gamma hyperparameters $(a_\eta, b_\eta)$ at inference time
must match those used at training time; the released
$K{=}4$ EM-warmup checkpoint uses
$(a_\eta, b_\eta) = (2, 2)$. The reduction $K_r = 1$ with
$\eta = 1$ is the canonical evaluation of that checkpoint, since
the stored per-site $\eta$ posteriors all collapsed to $1.0$
during EM warmup; for retrained checkpoints with a non-trivial
$\eta$ posterior, $K_r > 1$ is required.

\paragraph{$t = 0$ reduction identities.}
At $t = 0$ both per-block primitives reduce to the appropriate
marginal stationary:
\begin{align}
\label{eq:p-singlet-t0}
P_{\mathrm{singlet}}(a, a;\, 0)
  \;&=\; \sum_c \pi_c \, \pi^{(c)}(a) \;\equiv\; \pi_{\mathrm{marg}}(a), \\
\label{eq:p-doublet-t0}
P_{\mathrm{doublet}}\bigl((a, c), (a, c);\, 0\bigr)
  \;&=\; \sum_{c_1, c_2} \pi_c(c_1) \, \pi_c(c_2) \,
        \pi_{\mathrm{joint}}^{c_1, c_2}(a, c)
        \;\equiv\; \pi_{\mathrm{marg, pair}}(a, c).
\end{align}
These follow from $P_c(a \to b;\, 0) = \delta_{ab}$ and from the
joint stationary $\pi_{\mathrm{joint}}^{c_1, c_2}$ being a
fixed point of $P_{\mathrm{joint}}^{c_1, c_2}(\cdot;\, 0)$.

\paragraph{The boost tensor.}
The four-residue Potts \emph{boost} entering equation~\ref{eq:tkfdp-joint-marginal}
through the doublet factor is the multiplicative deviation of the
doublet emission from the product of two singlet emissions:
\begin{equation}
\label{eq:M-boost}
M\bigl((a, c), (b, d); t\bigr)
  \;\equiv\;
  \frac{P_{\mathrm{doublet}}\bigl((a, c), (b, d); t\bigr)}
       {P_{\mathrm{singlet}}(a, b; t) \cdot P_{\mathrm{singlet}}(c, d; t)}.
\end{equation}
Indexing convention as above: $(a, b)$ is the left endpoint, $(c, d)$
the right endpoint. By construction $M \equiv 1$ when every Potts
atom is zero (so $\pi_{\mathrm{joint}}^{c_1, c_2}$ factorizes into
$\pi^{(c_1)} \otimes \pi^{(c_2)}$ and $P_{\mathrm{joint}}^{c_1, c_2}$
factorizes into the two single-site $P_{c_i}$ factors). The boost
tensor $M$ is the object that the bounded-edge augmented Pair HMM
methods of Appendix~\ref{app:phmm} fold multiplicatively into the
Match emissions to score doublets relative to singletons. The MCMC
sampler operates on the joint $(A, E)$ directly via
equation~\ref{eq:tkfdp-joint-marginal}, with class assignments and
rate multipliers integrated analytically as above (\emph{not}
sampled).

\subsubsection*{Sequence annealing with coevolutionary scoring}
\label{sec:coupled-annealing}

The joint-marginal of equation~\ref{eq:tkfdp-joint-marginal}
implies a per-Match-cell posterior $Q'_{ij}$ that the augmented
Pair HMM machinery of Appendix~\ref{app:phmm} computes either
exactly under bounded-edge truncation or stochastically under the
infinite-Pair-HMM MCMC. An alternative routing of the same
coupling signal applies the boost \emph{during} assembly: the
greedy column-merging step inside the sequence annealer is extended
to score pairs of merges jointly. With pair-HMM input, the
canonical edge score is
\begin{equation}
\label{eq:tgf-edge}
w(e) \;=\; \frac{2 \, Q_{ij}^{(X, Y)}}{g_i^{(X, Y)} + g_j^{(Y, X)}},
\qquad e = \bigl(\text{column of } X_i,\ \text{column of } Y_j\bigr),
\end{equation}
where $g$ are the per-residue gap posteriors. The coupled-pair
extension admits joint candidates consisting of pairs of merges
scored as
\begin{equation}
\label{eq:coupled-score}
W(e_1, e_2) \;=\; \log Q_{ij}^{(X, Y)} + \log Q_{i'j'}^{(X, Y)}
  + \log M\bigl((X_i, X_{i'}), (Y_j, Y_{j'}); t\bigr),
\end{equation}
with $M$ the boost tensor of equation~\ref{eq:M-boost} evaluated
at the observed amino acids. The first two terms are the baseline
log-posteriors, and the third is the four-residue Potts
contribution; the latter is identically zero when every Potts
atom is zero.

\paragraph{Greedy schedule and conflict resolution.} When a coupled
candidate $(e_1, e_2)$ is popped from the queue, conflicts are
resolved as follows. If both component merges are still admissible
(no same-sequence conflict, no DAG cycle), both are committed in a
single step and the boost-driven joint commit is recorded. If $e_1$
is admissible but $e_2$ has been invalidated by an intervening
merge, the candidate degrades to a single-edge merge with score
$\log Q_{ij}$ and is re-inserted into the queue at that priority.
If neither component is admissible, the candidate is dropped.

\paragraph{Cost analysis.} Under two practical prunings (threshold
$q_{\min} = 0.1$ on the single-edge posterior and threshold
$\mu_{\min} = 0.1$ nat on $|\log M|$), the per-step cost is
$O(L^2)$ amortized and the total cost is $O(L^3)$ in the worst
case, comparable to the $O(L^2 \log L)$ baseline assembler.

\paragraph{Augmented-PHMM versus greedy-pairwise tradeoff.}
The augmented-Pair-HMM and infinite-Pair-HMM methods of
Appendix~\ref{app:phmm} produce a single corrected posterior
matrix $Q'$ that any downstream assembler can consume. The
coupled-pair extension above scores strongly-coupled pairs jointly
on their native four-residue domain inside the assembler, but
commits greedily and is therefore vulnerable to local optima where
an early commit on a coupled pair locks out a globally better
arrangement. The two routes consume the same boost tensor $M$ of
equation~\ref{eq:M-boost} but apply it at different points in the
pipeline; the augmented-PHMM route is the principled default for
TKF-DP postprocessing in \S\ref{app:phmm}.
